---
title: "Identification of SSTs_report"
author: "Lisa Marie Milchsack"
date: "10 Juni 2019"
output:
  html_document: default
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Installing required packages
```{r include=FALSE} 
#i think this is obsolete in the end, we can keep it, as long as there is confusion about what packages are needed
install.packages("reshape")
install.packages("wesanderson")
install.packages("tidyverse")
install.packages("factoextra")
install.packages("data.table")
library(tidyverse)
library(cluster)
library(factoextra)
library("reshape")
library("wesanderson")
library(data.table)
```

# Abstract
(...)

### Introduction
## Revision: Breast Cancer, a challenging disease

> "The heterogeneity of breast cancers makes them fascinating and challenging to diagnose and treat." 

According to the latest reports of breast cancer prevalence, the current state is very serious:
  + Breast cancer ranks #2 in worldwide cancer case numbers (Hutchinson 2010)
  + In Germany there are 70,000 new cases annually. (Barnes, Kraywinkel et al. 2016)
  + This makes ~ 30% of all reported cases of tumor diseases.(Robert-Koch-Institut 2017)

# Driver Mutation
In our project, we first focus on driver mutations, which can be imagined as the main "force" in tumorigenesis. Certain genes induce/improve tumor growth as consequence of mutation, such as:
  + Providing tumor cells with Growth advantage
  + Leading to positive selection in microenvironment
Furthermore, driver mutations are also often required for tumor maintenance. (Stratton, Campbell et al. 2009)
  
## Project design: Our selection of possible driver mutation

+ From literature, we selected possible driver mutations based on three main characteristics:
  1. overexpressed in breast cancer
  2. observed in a relevant percentage of clinical cases
  3. part of processes/pathways of tumorgenesis
  
+ Finally, we selected:

Gene Name | Overexpression rate | Role in Tumorgenesis                   | Sources                                                     |
----------|---------------------|----------------------------------------|-------------------------------------------------------------|
CCND      | ~ 50%               |Cell cycle regulator                    |(Arnold and Papanikolaou 2005)                               |
ERBB2     | ~ 30%               |Encodes endothelial growth factor       |(Slamon, Godolphin et al. 1989)                              |
MCY       | ~ 30 - 50%          |Cell cycle and apoptosis regulator      |(Gabay, Li et al. , Xu, Chen et al. 2010)                    |
PARP      | ~ 45%               |Regulates differentiation/proliferation |(Siraj, Pratheeshkumar et al. 2018)                          |
PIK3CA    | ~ 30 - 48%          |Interaction with AKT and mTOR pathway   |(Aleskandarany, Rakha et al. 2010, Shimoi, Hamada et al.2018)|

## Revision: The concept of Second Site Targets (SSTs)
Targeting and eliminating driver mutations as a part of tumor treatment is a difficult procedure, since those genes are also essential for survival of non-mutated cells. A differentiation between normal and tumor cell is not possible, thus targeting leads to adverse effects. Plus, high dosages need to be applied. #CITATION
A promising approach is targeting so-called Second Site Targets (SSTs), which are gene mutations interacting genetically with driver mutations to increase cell viability. # CITATION
If knocking out those genes leads to cell lethality, it may lead to new treatment options. 

Our central question for our project is thereby:
> Which __second-site targets__ interact genetically with __driver mutations__ to promote cell viability and proliferation in breast cell cancers?

# Main

## Data cleanup

Before we start our statistical part, we have to load the data correctly:
```{r include=FALSE}
#Importing the libs
library(ggplot2)
library(reshape2)
library(data.table)
library(cluster)
library(caTools) 
library(tidyverse)
library(factoextra)

#Setting a sys-path
root.dir = dirname(rstudioapi::getSourceEditorContext()$path)

#Importing the data
data = readRDS(paste0(root.dir, "/DepMap19Q1_allData.RDS")) #read in the data (with paste0 you can concatinate strings; here some paths this makes your code very flexible)
```

To get a feeling for the dataset we will have a look on data structure 

```{r include=FALSE}
length(data) #get some info about the data (the dimensions here)
lapply(data, dim)#get some impression about the data
view(data$mutation) #closer look because data$mutation returns NULL in the command above
lapply(data, function(a) head(a)) #look at the data 
```

We find a dataset containing 6 smaller datasets. Five of them are large dataframes while a sixth one is a list of dataframes. For the extraction of this sixth variable we first separated it fom the others and gave it a new strucure, which is easier to handle. 

```{r}
#Extract the mutation data
mut <- data$mutation #pick the mutation data (this has a specific format and neets to be treated separately)
'%!in%' <- function(x,y)!('%in%'(x,y)) #define an operator that will only pick the data that is NOT defined in the list; 
dt_new <- lapply(which(names(data) %!in% "mutation"), function(a) data[[a]]) #extract only non-mutation data
names(dt_new) <- names(data)[which(names(data) %!in% "mutation")] #rename the data with the original names
length(dt_new) #look if it worked (now dim is 5 and NOT 6 (because mutation was removed))
sample_case = c("Breast Cancer") #gives us the opportunity to analyse other groups of cells
samples = data$annotation$DepMap_ID[which(data$annotation$Primary.Disease == sample_case)]
ids = which(names(mut) %in% samples)
dat = lapply(ids, function(a) mut[[a]])
```

We observe a large variety of tumor diseases beeing recorded by this dataset. Because we want to focus on breast cancer we will extract the data contributing to this harmful ilness. We identified 28 cell lines, fulfilling this condition.

```{r}
#finding all breast cancer cell lines
BCCL_Numbers <- which(allDepMapData$annotation[4]== "Breast Cancer")
BCCL_Annotation <- subset(allDepMapData$annotation, Primary.Disease == "Breast Cancer")
```
From these cell lines we got expression values, copynumbers, and knockdown scores in cell viability and probability of cell survival as described in the introduction. To make data more clear, we constructed dataframes for every *sort of data* and ordered the genes alphabetically. 
```{r}
#summary(BCCL_Annotation)
# copy expression matrix to new dataframes
df1 <- allDepMapData$expression
df2 <- allDepMapData$copynumber
df3 <- allDepMapData$kd.ceres
df4 <- allDepMapData$kd.prob
#change column names to numbers for easier handling
names(df1) <- c(1:544)
names(df2) <- c(1:544)
names(df3) <- c(1:544)
names(df4) <- c(1:544)
#extract BCCL as dataframe
BCCL_Expression <- as.data.frame( df1[, c(BCCL_Numbers)])
BCCl_Copynumber <- as.data.frame( df2[, c(BCCL_Numbers)])
BCCL_kd.ceres <- as.data.frame( df3[, c(BCCL_Numbers)])
BCCL_kd.prob <- as.data.frame( df4[, c(BCCL_Numbers)])
dim(BCCL_Expression)
dim(BCCl_Copynumber)
dim(BCCL_kd.ceres)
dim(BCCL_kd.prob)
#reorder rows alphabetically
BCCL_Expression <- BCCL_Expression[order(rownames(BCCL_Expression)), ]
BCCl_Copynumber <- BCCl_Copynumber[order(rownames(BCCl_Copynumber)), ]
BCCL_kd.ceres <- BCCL_kd.ceres[order(rownames(BCCL_kd.ceres)), ]
BCCL_kd.prob <- BCCL_kd.prob[order(rownames(BCCL_kd.prob)), ]
#subset Mutation matrix in BCCL
BCCL_Mutation <- allDepMapData$mutation[c(BCCL_Numbers)]
```
As next step we made some experiments, to get a feeling for data distribution. 
We made a heatmap of expression values, compared single genes for knockdown and copynumber scores and found relatively equally distibuted expression values and CERES Scores. We also tried to gain a first look on driver mutations, known from literature.
*muss noch sch?n gemacht werden :)*
```{r echo=TRUE}
tBCCL_Expression <- t(BCCL_Expression)
boxplot(tBCCL_Expression[, c("MYC", "PTEN", "TP53", "PIK3CA", "GATA3", "BRCA1", "BRCA2", "RB1", "GANAB", "PRKCSH", "RNF43")], xlab = "Suggested driver mutations", ylab ="Transcripts per million")
```
We already discarded some of this genes, mostly because they are known as tumor suppressor genes. In our approach we looked for targets, which show similar characteristics in cell viability after a knockdown. A knockdown of a tumor supressor gene, even if it is mutated, will rarely have a negative effect on cell viability. This makes genes associated with tumor supressor genes useless for an inhibitory drug and therefore useless for our analysis. 

### Generation of the mutImpact matrix

Vorschlag: Every mutation has an impact on cell viability. In our analysis we have to distinguish not only between single mutations, we also have to evaluate different mutations in our cell lines "working together" and making every cell line different in its chcaracteristics. In order to gain an overview of these characteristics, we will generate a matrix, showing the effect on viability, caused by a single mutation, in every cell line. 
In order to gain an overview of the effects on viability of each mutation present in one cell line the mutImpact matrix was generated.

Therefore, the BCCL_kd.ceres matrix containing CERES scores for all genes in breast cancer cell lines was used as a template and the cell line IDs were added as column names creating a pre_mutImpact matrix. An empty matrix was generated comprising 17634 rows and 28 columns matching the number of genes per cell line and the number of breast cancer cell lines, respectively. 

Subsequently, CERES scores of genes which were not mutated in a given cell line were replaced with "NA" by implementing an if statement. This statement included the enquiry whether the gene ID is present in the list of mutated genes of a considered cell line as and in case the condition was met the CERES score was maintained. If "FALSE" was returned the CERES score of the gene was replaced with "NA" for the considered cell line.

```{r, warning=FALSE}
#Generate pre_mutImpact matrix containing all CERES values and with IDs as colnames
pre_mutImpact <- BCCL_kd.ceres
colnames(pre_mutImpact)<-tPatients_ID

#generate empty mutImpact matrix
mutImpact <- matrix(, nrow = 17634, ncol = 28)
colnames(mutImpact)<-colnames(pre_mutImpact)
rownames(mutImpact)<-rownames(pre_mutImpact)

#fill mutImpact matrix with values
for (j in 1:ncol(pre_mutImpact)){
  lineID <- colnames(pre_mutImpact)[j] #select a column-name = cell line
  for (i in 1:nrow(pre_mutImpact)){
    GOI <- rownames(pre_mutImpact)[i] #select a gene
    if (GOI %in% BCCL_Mutation[[lineID]]$Hugo_Symbol){
      mutImpact[i, j] <- pre_mutImpact[i, j]
    } else {
      mutImpact[i, j] <- NA
    }#replace CERES value with NA in case the gene is not mutated
  }
}

#have a look at the mutImpact matrtix
head(mutImpact)
```

### Identification of driver mutations using k means clustering of CERES scores 

In order to verify our selection of driver mutations from literature research, k means clustering of CERES scores was performed. Suitable driver mutations were expected to be found in the same cluster based on scaled CERES scores.

```{r kmeans ceres}

#Loading mutImpact matrix
pre_mutImpact <- BCCL_kd.ceres
colnames(pre_mutImpact)<-tPatients_ID
mutImpact <- matrix( , nrow = 17634, ncol = 28)
colnames(mutImpact)<-colnames(pre_mutImpact)
rownames(mutImpact)<-rownames(pre_mutImpact)
for (j in 1:ncol(pre_mutImpact)){
  lineID <- colnames(pre_mutImpact)[j] #select a column-name aka cell line
  for (i in 1:nrow(pre_mutImpact)){
    GOI <- rownames(pre_mutImpact)[i] #select a gene
    if (GOI %in% BCCL_Mutation[[lineID]]$Hugo_Symbol){
      mutImpact[i, j] <- pre_mutImpact[i, j]
    } else {
      mutImpact[i, j] <- NA
    }#replace CERES value with NA in case the gene is not mutated
  }
}
```

K means clustering was performed with all genes mutated at least once out of all 28 cell line samples and whose CERES score was equal to or lower than zero. For this, all genes with only NA values, hence not one mutation in any cell sample, were removed. A new dataframe `BCCL_kd.ceres_2`was generated containing all 28 CERES scores for each gene, with gene names in the first column. Gene names in the first column of the data frame `BCCL_kd.ceres_2`were compared to the rownames of `mutImpact_c` to generate the BCCL_kd.ceres_3 dataframe containing CERES scores for all breast cell cancer lines for all genes mutated at least once. Subsequently, the mean CERES score for each gene was computed using the rowmeans function and saved in the last column. For this, the column `rownameskdc2`had to be removed. Finally, all genes with CERES scores > 0 were deleted. The final BCCL_kd.ceres_3 dataframe now contains 4934 genes mutated at least once in the 28 breast cancer cell lines and with CERES scores equal to or below 0. 

```{r}
# selecting only genes mutated at least once and with Ceres >= 0
mutImpact_c = mutImpact[rowSums(is.na(mutImpact)) != ncol(mutImpact),]# remove all genes withall NA values
rownamesmutImpc = rownames(mutImpact_c)
mutImpact_c = cbind(rownamesmutImpc, mutImpact_c[,2:ncol(mutImpact_c)]) # insert new column with row names
dim(mutImpact_c) # check gene reduction 
rownameskdc2 = rownames(BCCL_kd.ceres) 

BCCL_kd.ceres_2 = cbind(rownameskdc2, BCCL_kd.ceres[,2:ncol(BCCL_kd.ceres)]) # insert new column with row names
BCCL_kd.ceres_3 = BCCL_kd.ceres_2[BCCL_kd.ceres_2[,"rownameskdc2"] %in%  mutImpact_c[,"rownamesmutImpc"],][,-c(1)]
BCCL_kd.ceres_3 = cbind(BCCL_kd.ceres_3, rowMeans(BCCL_kd.ceres_3))[rowMeans(cbind(BCCL_kd.ceres_3, rowMeans(BCCL_kd.ceres_3)))<= 0,] # deletion of genes with CERES > 0
dim(BCCL_kd.ceres_3)
```

In preparation of k means clustering, the column containing average CERES scores for each gene was removed (new dataframe `BCCL_kd.ceresKS`) and the optimal number of k clusters identified. For each column, so cell samples, of the dataframe the data values were scaled. For cluster numbers between 2 und 15 the within sum of squares for each cell sample was calculated and stored as a seperate dataframe. To ensure that the same result is obtained each time the code is run the set.seed(1234) function was used.  

```{r}
#determining optimal cluster number k
BCCL_kd.ceresKS = BCCL_kd.ceres_3[,-c(28)] # remove column of average CERES value column 

CERES_Optimal_K <- function(BCCL_kd.ceresKS, specifier) {
  output <- lapply(1:ncol(BCCL_kd.ceresKS), function(a){
   df <- scale(BCCL_kd.ceresKS[,a]) # pick one column of input data and scale 
   wss <-(nrow(df-1))*sum(apply(df,2,var)) #compute wss
   for (i in 2:15){ # for k between 2 and 15
     set.seed(1234)
     wss[i] <- sum(kmeans(df, centers = i)$withinss)
     } 
   return(wss)
  })
names (output) <- colnames(BCCL_kd.ceresKS) # rename the output 
return(output)
}

BCCL_kd.ceres_optKS <- CERES_Optimal_K(BCCL_kd.ceresKS, "CERES Optimal Clusters k") 
```

To enable the plotting of within sum square (WSS) values for clusters from 2 to 15 for each breast cancer cell, each vector was individually picked and formatted into a dataframe. Cell sample name was then added as a label in a new column "Cell_Sample" and the cluster numbers from 1 to 15 in the column "OptimalK".
The output of this lapply function (`optimalKprocessedData`), a list of 28 dataframes, was then combined into one and columns renamed accordingly. The ggplot function was then used to plot all data stored in `optimalKprocessedData` in one plot. 

```{r}
# plotting wss function 
optimalKprocessedData <- lapply(seq_along(BCCL_kd.ceres_optKS), function(a){
  dtPicker <- as.data.frame(BCCL_kd.ceres_optKS[[a]]) # one vector picked and formatted into dataframe
  dtPicker$Cell_Sample <- names(BCCL_kd.ceres_optKS)[a] #sample added as label
  dtPicker$OptimalK <- 1:nrow(dtPicker) 
  return(dtPicker)
})

optimalKprocessedData <- as.data.frame(rbindlist(optimalKprocessedData)) # output into dataframe
colnames(optimalKprocessedData) <- c("WSS", "Cell_Sample", "OptimalK")

# using ggplot to plot output 
ggplot(data = optimalKprocessedData, aes(x=OptimalK, y=WSS)) +
  geom_line(aes(color=Cell_Sample)) +
  labs(title = 'Optimal number of clusters k', x = 'Number of clusters', y = 'WSS') +
  theme_minimal() +
           theme(legend.position ='bottom',
                  plot.title = element_text(hjust = 0.5),
                  axis.text.x = element_text(angle = 0, vjust = 0, hjust=0.5),
                  legend.title= element_blank(),
                  axis.title.x = element_blank(),
                  strip.text.y = element_text(angle = 0))
```
From this plot it is evident that the optimal cluster number, when applying the elbow method, lies between 2 and 4. To verify this result and to determine the optimal cluster number, an alternative plotting method calculating the average WSS for cluster numbers between 2 and 15 for all 28 cell samples was performed. 

```{r} 
# Verification of optimal number of clusters k 

km_CERES= kmeans(x = BCCL_kd.ceres_3, centers = 2, nstart = 10)
table(km_CERES$cluster)
wss = sapply(2:15, function(k) {
    kmeans(x = BCCL_kd.ceres_3, centers = k)$tot.withinss
})
plot(2:15, wss, type = "b", pch = 19, xlab = "Number of clusters K", ylab = "Total within-clusters sum of squares") 
```

The optimal cluster number k was identified to be 4. Thus k means clustering of the data in BCCL_kd.ceres_3 was performed with 10 iterations. The following cluster plot was generated.

```{r}
km_CERES= kmeans(x = BCCL_kd.ceres_3, centers = 4, nstart = 10)
table(km_CERES$cluster)
fviz_cluster(km_CERES, data=BCCL_kd.ceres_3, labelsize = 6) # plotting clusters
```

In order to verify our selection of driver mutations from literature, the cluster number of each driver mutation was identified. All selected driver mutations from literature,"ERBB2", "MYCBP", "PARP10" and "PIK3CA" were colocalized in cluster 3, hence with similar CERES scores, with the exception of "CCND" and "ERBBIP". While "CCND" was removed in the filtering process and no longer found in`BCCL_kd.ceres_3`, "ERBB2IP" was found in cluster 2. For these reasons the mentioned two mutations were excluded from further analysis, to leave only "ERBB2", "MYCBP", "PARP10" and "PIK3CA" as selected driver mutations for further analysis. 

``` {r}
km_CERES$cluster["ERBB2"] #obtaining cluster number for each driver mutation from literature 
km_CERES$cluster["ERBB2IP"]
km_CERES$cluster["MYCBP"]
km_CERES$cluster["PARP10"]
km_CERES$cluster["PIK3CA"] 
```

Since our approach defined SSTs as genes expected to interact synergistcally with the corresponding driver mutation in promoting cell viability, similar CERES scores were expected. Thus all potential SSTs candidates were expected to also be found in cluster 3. A dataframe `data_clus_3`was subsequently generated to contain all 1148 genes colocalized with driver mutations in cluster 3. 

```{r}
data_cluster = km_CERES$cluster
rownames(data_cluster) = rownames(BCCL_kd.ceres_3)
data_clus_3 = BCCL_kd.ceres_3[data_cluster == 3,] #dataframe with all 1148 genes in same cluster as driver mutations 
```

A further dataframe `data_clus_3e` was subsequently generated labelling the four driver mutations in a separate column, by implementing an "IF" loop. If the rowname of `data_clus_3e`was found in a previously created vector containing driver mutations names (`drivermut`), it was labelled as "driver mutation".Otherwise genes were labelled as "no driver mutation".

```{r Labeling driver mutations}

drivermut =  c("ERBB2","MYCBP", "PARP10", "PIK3CA") # create vector with driver mutation names

data_clus_3e = data_clus_3
data_clus_3e["Driver mutation"] <- sapply(1:nrow(data_clus_3), function(j){
     if(rownames(data_clus_3e)[j] %in% drivermut){
         data_clus_3e[j,1] = "driver mutation"
     } else {
         data_clus_3e[j,1] = "no driver mutation"
     }
  }) # labeling driver mutations with "driver mutation" in data_clus_3e 
```

### Data visualisation of CERES and TPM values for the four selected driver mutations 

CERES scores and expression as TPM values for the selected 4 driver mutations were visualised as box and whisker plots. Following the generation of the `GOI_X` dataframe containing either CERES or TPM values for the four mutations, the data was reshaped to contain two columns. One column being "variable" with the names of driver mutations, the other "value" with the corresponding CERES or TPM value. The ggplot function, part of the ggplot2 package, was used to plot the following box an whisker plots. 

```{r data visualisation, fig.align = center}

# CERES_boxplot
GOI= BCCL_kd.ceres[c("ERBB2","MYCBP", "PARP10", "PIK3CA"),]
CERES_driver_mut= melt(GOI)
CERES_driver_mut$variable=rep(c("ERBB2","MYCBP", "PARP10", "PIK3CA"), ncol(GOI))

CERES_boxplot = ggplot(CERES_driver_mut, aes(x = variable, y = value)) +
       geom_boxplot(aes(fill=variable), outlier.alpha = 0.7,
                     outlier.colour = "grey", outlier.shape = 20, outlier.size = 2) +
        labs(title = 'CERES scores of selected driver mutations', x = 'Selected driver   mutations', y = 'CERES scores') +
            theme_minimal() +
            geom_jitter(width = 0.2) +
            scale_fill_manual(values=wes_palette(n=4, name="Royal2")) +
            theme(legend.position ='none',
                  plot.title = element_text(hjust = 0.4),
                  axis.text.x = element_text(angle = 0, vjust = 0, hjust=0.5),
                  legend.title= element_blank(),
                  axis.title.x = element_blank(),
                  strip.text.y = element_text(angle = 0))
CERES_boxplot

  
#Expression_boxplot
GOI_Exp= BCCL_Expression[c("ERBB2","MYCBP", "PARP10", "PIK3CA"),]
Expression_driver_mut= melt(GOI_Exp)
Expression_driver_mut$variable=rep(c("ERBB2","MYCBP", "PARP10", "PIK3CA"), ncol(GOI_Exp))

Expression_boxplot = ggplot(Expression_driver_mut, aes(x = variable, y = value)) +
       geom_boxplot(aes(fill=variable), outlier.alpha = 0.7,
                     outlier.colour = "grey", outlier.shape = 20, outlier.size = 2) +
        labs(title = 'Transcripts per million (TPM) of selected driver mutations', x = 'Selected driver mutations', y = 'Transcripts per million (TPM)') +
            theme_minimal() +
            geom_jitter(width = 0.2) +
            scale_fill_manual(values=wes_palette(n=4, name="Royal2")) +
            theme(legend.position ='none',
                  plot.title = element_text(hjust = 0.4),
                  axis.text.x = element_text(angle = 0, vjust = 0, hjust=0.5),
                  legend.title= element_blank(),
                  axis.title.x = element_blank(),
                  strip.text.y = element_text(angle = 0))
Expression_boxplot
```

The box and whisker plot of CERES scores reveals highest essentiallity of the MYCBP driver mutation for cell survival, through the lowest average CERES scores. This is followed by PIK3CA and PARP10 mutations. While ERBB2 has the highest average CERES scores and hence lowest impact on cell proliferation, this mutation exhibits the largest spread of values across 28 breast cell cancer lines, reaching the overall minimum value of -1. The interquartile range of CERES scores for both MYCBP and PARP10, are lowest, indicating a consistent impact of gene knockouts on cell sample survival.To be noted also is the large interquartile range of the PIK3CA mutation, with data points found in two clusters, one around -0.15 and the other bweetn -0.5 and -0.9. 

While the ERBB2 driver mutation was found to have on average the lowest essentiallity for cell survival, average TPM values are shown to be highest, suggesting overexpression. Once again TPM values for this mutation have the largest spread, reaching a maximum over 11.5, implying greatest variation overall in data for the ERBB2 driver mutation. Like for the CERES scores, both MYCBP and PIK3CA have a small interquartile range, while ranges for PARP10 are markedly larger. Further, it is interesting to note despite two CERES score data clusters for PIK3CA, TPM values show little variation around the mean. 

### Spotting genetic interactions: the Wilcoxon test

Genetic interactions of driver mutations with other genes were examined based on a Wilcoxon signed rank test. As SSTs are expected to interact synergistcally with the corresponding driver mutation in promoting cell viability similar CERES scores are suggested to infer the presence of gene interactions. Thus, high p-values are anticipated for such cooperating gene pairs when performing the Wilcoxon signed rank test. The Wilcoxon signed rank test was chosed rather than a t-test due to non-normal distribution of the data and the requirement of a paired test to compare the data of two genes per cell line. Asuming that the ranks of the CERES scores of the driver mutation and their SSTs are equal SST candidates were chosen by the means of the highst p-values (accepting the H0 hypothesis).

For the purpose of conducting the Wilcoxon test the data from which the CERES scores for the test are taken was prepared. Therefore, a vector containg the identified driver mutations was generated. Furthermore, a reduced mutImpact matrix `mutImpact_r` was created from `mutImpact`which only includ genes which are mutated and do not have `NA`values only. Based on `mutImpact_r`the `BCCL_kd.ceres` matrix was reduced resutling in the `mutImpact_kd.ceres` matrix containing the CERES scores of all genes in all cell lines if the gene is mutated in one cell line at minimum. 

```{r}
#preparation of the data
driver_mut <- c("ERBB2","MYCBP", "PARP10", "PIK3CA") #enter driver mutations
mutImpact_r <- mutImpact[rowSums(is.na(mutImpact)) != ncol(mutImpact), ] #_r=reduced: get rid of all the ONLY NA rows (this will save computation)
mutImpact_kd.ceres <- BCCL_kd.ceres[rownames(BCCL_kd.ceres) %in% rownames(mutImpact_r),] # create matrix containing CERES scores of genes which are mutated once at minimum
colnames(mutImpact_kd.ceres) <- tPatients_ID
```

A Wilcoxon signed rank test was performed for every driver mutation with all other mutated genes and the data was stored in a list comprising one data frame with the obtained p-values for each driver mutation. 

Thereupon, a loop was installed generating the output data frame for each driver mutation. At first the driver mutation to be examined was selected and the corresponding CERES values retrieved from `mutImpact_kd.ceres` being saved in the temporary variable `driverMutData`. Consequently, a reference gene was chosen and the CERES scores of the reference gene was stored in another temporary vector `refGeneData`. In case the selected reference gene was not the driver mutation itself the Wilcoxon test was performed. The p-value of each gene pair was saved in the data frame of the considered driver mutation.

```{r}
#performance of the Wilcoxon test
testData <- lapply(seq_along(driver_mut), function(a) {
  driverMutPicker <- driver_mut[a] #pick a driver mutation
  driverMutData <- mutImpact_kd.ceres[driverMutPicker,] #get CERES scores of driver mutation
  
  outputData <- sapply(1:nrow(mutImpact_kd.ceres), function(b) {
    refGeneData <- mutImpact_kd.ceres[b,] #get CERES scores of reference gene (potential SST)
    
    if (rownames(refGeneData) != driverMutPicker) {
      out <- wilcox.test(as.numeric(driverMutData), as.numeric(refGeneData), paired = TRUE)$p.value #get p-value of wilcoxon signed-rank test
      out <- as.data.frame(out, rownames(mutImpact_r)[b]) 
      return(out)
    }
  })
  test = do.call(rbind, outputData)
  rownames(test) <- rownames(mutImpact_r)[which(rownames(mutImpact_r) != driverMutPicker)]
  return(test)
})
names(testData) <- driver_mut

lapply(testData, function(a) head(a)) #look at the data
```

As the null hypothesis claims that the CEREs scores of the driver mutation and the reference gene is are equal, low p-values indicate that there is no relation between the CERES scores of the driver mutation and the reference gene whereas high p-values may allude to genetic interaction of the driver mutation with an SST.
The obtained p-values were distributed as shown below. Most strinkingly, the first quantile was set at a p-value of 0.0 for each driver mutation indicating that there are many genes with CERES scores which are not relatable to the CERES scores of the driver mutations. However, the maximum p-value of each test constituted 1.0 suggesting that there are also genes with similar CERES scores as the driver mutations which are supposed to be the SSTs. 

```{r}
#get summary of distribution
summary(testData$ERBB2)
summary(testData$MYCBP)
summary(testData$PARP10)
summary(testData$PIK3CA)
```

Based on the results of the Wilcoxon test potential SSTs were chosen by means of the highest p-values. Thus, the list `SST_Data`consisting of one data frame per driver mutation was created based on the test results given in the `test_data` list. Therefore, an `lapply`command was used filtering each data frames of `test_data` by selecting the reference genes which were found to have the most similar CERES scores as the driver mutation, indicated by high p-values. In this course, the rownames of the data frames had to be transferred to an extra column before selection and transferred back afterwards as otherwise the rownames would have been lost during the filtering process. Unequal numbers of SST candidates were found due to similar test restults for several reference genes.

```{r}
#select 15 most promising SST candidates based on highest p-values - libraries needed: dplyr, tibble (included in "tidyverse")

SST_Data <- lapply(seq_along(testData), function(a) {
  data <- as.data.frame(testData[[a]], col.names = names(testData[[a]])) #create new data frame containing testData for each SST
  colnames(data) = driver_mut[a] #rename columns
  data <- data %>%rownames_to_column() %>% top_n(15, data[,1]) %>% column_to_rownames() #select 15 genes with highest p-value
  return(data)
})
names(SST_Data) <- driver_mut
lapply(SST_Data, function(a) print(a)) #have a look at the SSTs

#have a look at the SST candidates
head(SST_Data)
```

## Verification of SST candidates by assigned k means cluster number 

After selection of the SST candidates stored in `SST_cand_x`,loops were installed to verify the coexistance of SST candidates and driver mutations in the same k means cluster. An assignment to the same cluster would confirm a possible cooperation of gene pairs due to similar CERES scores. For this, new dataframes were created  before a separate loop was installed for each driver mutation. By comparing each rowname of the `SST_cand_X` dataframes to all possible  `data_clus_3e` rownames, a coexistance of genes in the 3rd k means cluster was identified, leading to the labelling of `SST X`  in a new separate column. Otherwise no match lead to the `No SST`entry. In order to combine all four dataframes with the `rbind`command the first columns were all renamed to "p_values". FInally all confirmed SSTs were ordered alphabetically, with no gene identified twice for a different driver mutation. 

```{r}
SST_ERBB2 <- SST_cand_ERBB2 #creating new dataframes
SST_MYCBP <- SST_cand_MYCBP
SST_PARP10 <- SST_cand_PARP10
SST_PIK3CA <- SST_cand_PIK3CA

#labelling of SST candidates with "SST" or "No SST" depending on if in k means cluster 3 or not

#PARP10 candidates
SST_PARP10["SST Cluster Check"] <- sapply(1:nrow(SST_cand_PARP10), function(a){
  if (rownames(SST_cand_PARP10) %in% rownames(data_clus_3e)) {
      SST_PARP10[a,1] ="SST PARP10"
  }else{
     SST_PARP10[a,1] = "No SST"
  }
})
SST_PARP10 <- SST_PARP10 %>% rename(p_values=PARP10)

#ERBB2 candidates
SST_ERBB2["SST Cluster Check"] <- sapply(1:nrow(SST_cand_ERBB2), function(a){
  if (rownames(SST_cand_ERBB2) %in% rownames(data_clus_3e)) {
      SST_ERBB2[a,1] ="SST ERBB2"
  }else{
     SST_PARP10[a,1] = "No SST"
  }
})
SST_ERBB2 <- SST_ERBB2 %>% rename(p_values= ERBB2)

#MYCBP candidates
SST_MYCBP["SST Cluster Check"] <- sapply(1:nrow(SST_cand_MYCBP), function(a){
  if (rownames(SST_cand_MYCBP) %in% rownames(data_clus_3e)) {
      SST_MYCBP[a,1] ="SST MYCBP"
  }else{
     SST_MYCBP[a,1] = "No SST"
  }
})
SST_MYCBP <- SST_MYCBP %>% rename(p_values= MYCBP)

#PIK3CA candidates
SST_PIK3CA["SST Cluster Check"] <- sapply(1:nrow(SST_cand_PIK3CA), function(a){
  if (rownames(SST_cand_PIK3CA) %in% rownames(data_clus_3e)) {
      SST_PIK3CA[a,1] ="SST PIK3CA"
  }else{
     SST_PIK3CA[a,1] = "No SST"
  }
})
SST_PIK3CA <- SST_PIK3CA %>% rename(p_values= PIK3CA)

#All approved SSTs in one dataframe with driver mutation label and p-value
SSTs <-  do.call("rbind", list(SST_ERBB2, SST_MYCBP, SST_PARP10, SST_PIK3CA)) 
SSTs <- SSTs[order(rownames(SSTs)),] # reorder in alphabetical order 

#no SSTs the same for different driver mutations 
```

## characterization of second-site targets

In order to characterize the SSTs, additional information on each SST was retrieved from the original data set and added to the data frames in `SST_Data`.

First, the type of mutation was examined for each SST and each cell line in which it was mutated. For this purpose a nested `lapply`construction was used. Initially, a data frame of one driver mutation was chosen and the new column `mutType` was generated by the second `lapply`loop. Thus, iterating the SST candidates the mutation status of each SST in each cell line was determined inspecting the presence of the SST in the list of mutated genes for each cell line. If the SST candidate was mutated in a given cell line its type of mutation (stored in the `BCCL_Mutation`list as `Variant_Classification`) was obtained and saved in a temporary vector to which all possible types of mutations were collected for each SST. Finally, this vector was concatenated as a string which was assigned to the `mutType` column.

```{r}
SST_Data <- lapply(seq_along(SST_Data), function(a) {
  dvPicker <- SST_Data[[a]]#pick a data frame of one driver mutation
  
  dvPicker$mutType <- sapply(seq_along(rownames(dvPicker)), function(b) {
    SSTPicker <- rownames(dvPicker)[b]#select an SST candidate
    
    mutTypeData <- c()
    for (i in 1:ncol(BCCL_kd.ceres)) { 
      if (SSTPicker %in% BCCL_Mutation[[i]]$Hugo_Symbol) { #check whether the SST mutation is present in a cell line
        out <- BCCL_Mutation[[i]]$Variant_Classification[a] #get type of mutation per SST and cell line
        mutTypeData <- c(mutTypeData, out) #store type of mutation in vector
      }
    }
    dvPicker[b, 2] <- paste(mutTypeData, collapse = ", ") #create a string of SST mutation types based on the vactor
  })
  return(dvPicker)
})
names(SST_Data) <- driver_mut
```

### Regression analysis to predict ceres scores of driver mutation
## Objective

  + For the regression model, we want to predict the CERES score of the driver mutation (dependent variable) based on the CERES score of the SST candidates (independent variables).
  + A multilinear regression will be designed:


# Step 1: Construct datasets for Regression
  + Multilinear regression, we 
    1. reform the BCCL CERES matrix
      + By transponing and format change, we obtain a matrix with colnames = Gene names and rownames = Cell line indices (equal to tPatients_ID)
    2. Collect the relevant the CERES scores for the mutations for the linear regression.
      + All columns with the corresponding names of the SST genes plus the driver mutation will be collected and copied into a new dataframe
      
```{r}
#reformat BCCL CERES matrix
BCCL_kd.ceres_df <- as.data.frame(t(BCCL_kd.ceres))
rownames(BCCL_kd.ceres_df) <- tPatients_ID
colnames(BCCL_kd.ceres_df) <- rownames(BCCL_kd.ceres)

#generate a dataframe containing data for each DV
linRegData_ERBB2 <- BCCL_kd.ceres_df %>% select(rownames(SST_Data$ERBB2), "ERBB2")
linRegData_MYCBP <- BCCL_kd.ceres_df %>% select(rownames(SST_Data$MYCBP), "MYCBP")
linRegData_PARP10 <- BCCL_kd.ceres_df %>% select(rownames(SST_Data$PARP10), "PARP10")
linRegData_PIK3CA <- BCCL_kd.ceres_df %>% select(rownames(SST_Data$PIK3CA), "PIK3CA")

data_regression <- list(linRegData_ERBB2, linRegData_MYCBP, linRegData_PARP10, linRegData_PIK3CA)
head(data_regression) #take a look at the data
```

#Step 2: qqplot
  + Check Normality with a qqPlot
    + Using the aes-function of the ggplt package for _aesthetic_ plots.
    
```{r}
create_qqplot <- function(a){  
  p <- ggplot(data_regression[[a]], aes(sample=data_regression[[a]][,ncol(data_regression[[a]])])) + stat_qq()+ stat_qq_line()
   p + labs(title = "QQ-Plot", subtitle = "Check if data is approx. normally distributed")
}
lapply(seq_along(data_regression), create_qqplot)
```
#Step 3: Multilinear Regression: Data split, Model calculation, Return correlation data for evaluation
To run the Multilinear Regression, it is advised to create a long function which
  1. Splits the dataset in 75:25 training:test data
    + Previous installation of the "caTools" package is necessary
  2. Conducts Training of Regression model with training data
    + predict Driver mutation CERES (y; dep.variable) from SST CERES scores (x1 ... xn; indep.variables)
  3. Runs Test Data on model for performance evaluation
  4. Calculates Spearman correlation to further asess model performance
  
```{r}
input_data <- data_regression

lapply(seq_along(input_data), function(a) {
  set.seed(123) #initialize the random numbers, better reproducable
  data <- input_data[[a]] #get the data
  colnames(data)[length(colnames(data))] <- "Predictor"
  
  split = sample.split(data[,ncol(data)], SplitRatio = 0.75) #split the dataset into 3/4 Training and 1/4 Testing dataset
  training_set = subset(data, split == TRUE) #use the labels to get the training data
  test_set = subset(data, split == FALSE)

  regressor = lm(formula = Predictor ~ ., 
                 data = training_set)  #predict CERES of Driver Mutation based on all (=.) the input variables (SSTs)
  
  y_pred = predict(regressor, newdata = test_set) #predict the Driver mut. CERES score based on the test data
  test_set$Prediction = y_pred #adding predictions to the dataset
  corVal <- cor.test(test_set$Predictor , test_set$Prediction, method = "spearman")#to judge model performance, calculate Spearman corr. between predicted and observed values
  return(corVal)
})
```

## Interpretation of Spearman correlation coefficient

# Revision: Meaning of the correlation coefficient rho
  + the rank-correlation coefficient can take values from +1 to -1
    + rho ~ 1: perfect positive association of ranks
    + rho ~ 0: no association of ranks
    + rho ~ -1: perfect negative association of ranks

# Revision: Adding a statistical parameter
  + To further interprete the p-value, we construct a H0 / H1 hypothesis:
    + H0: There is no association between the predicted and observed variable.
    + H1: There is an association between the predicted and observed variable.
  + Most importantly, it is not possible to assess the strength of the correlation based on the p-value. It rather judges the statistical test than the resulting value.
    + as a standard, we set the significance level to 0.05

# Objective
 + We will the compare this values for the different models
  + For a positive correaltion, two variables move in the same direction.
    + For a well-designed model, we expect the precicted and observed CERES scores being approx. even
  + Thus, optimal model design is achieved if:
      + the correlation coefficient is close to 1
      + the p-value < 0.05
    

# Results

We obtained the followwing p-values and correlation coefficients:

Model     | p-value     | rho      | 
----------|-------------|----------|
ERBB2     | 0.8397      | 0.1071   |
MCYBP     | 0.7825      |-0.1429   |
PARP      | 0.1095      |-0.6786   |
PIK3CA    | 0.7131      |-0.1786   |

# Discussion

  + ERBB2-model:
    + Shows the highest rho value of all four models, but ~ 0.1 is a rather weak positive relationship
    + The p-value leads to __acceptance of H0__.
    + The data is not even approx. normal distributed.
    
  + MYCBP-model:
    + The rho value leads to very weak negative correlation.
    + A p-value of ~0.78 leads to condifent acceptance of H0.
    + Data is approx. normal distributed.
    
  + PARP10-model:
    + The rho value of ~ -0.67 is a moderate negative relationship.
    + The p-value leads to acceptance of H0. The p-value is the lowest with ~ 0.11. But even if we would raise the confidence level to 0.10 (which would be a questionable move in statistics), the result wouldn´t be significant to accept H1.
    + Data is approx. normal distributed.
    
  + PIK3CA-model:
    + The rho value leads to very weak negative correlation.
    + A p-value of ~0.78  leads to condifent acceptance of H0.
    + The data is not even approx. normal distributed.

+ __All in all, all multilinear regression models failed to deliver a p-value of significant relevance. Thus, they should not be used for further predictions.__
+ The models for MYCBP and PIK3CA only showed weak, negative correlation, which is not useful for further work with this model.
+ The PARP10 model resulted in a moderate negative relationship, which is not of relevance for our objective.
+ The ERBB2 model did not deliver a significant p-value, but indeed contains at least a weak positive correlation.

### Literature
  + Aleskandarany, M. A., E. A. Rakha, M. A. H. Ahmed, D. G. Powe, E. C. Paish, R. D. Macmillan, I. O. Ellis and A. R. Green (2010). "PIK3CA expression in invasive breast cancer: a biomarker of poor prognosis." Breast Cancer Research and Treatment 122(1): 45-53.
  + Arnold, A. and A. Papanikolaou (2005). "Cyclin D1 in Breast Cancer Pathogenesis." Journal of Clinical Oncology 23(18): 4215-4224.
  + Barnes, B., K. Kraywinkel, E. Nowossadeck, I. Schönfeld, A. Starker, A. Wienecke and U. Wolf (2016). Bericht zum Krebsgeschehen in Deutschland 2016, Robert Koch-Institut.
  + Gabay, M., Y. Li and D. W. Felsher "MYC activation is a hallmark of cancer initiation and maintenance." Cold Spring Harbor perspectives in medicine 4(6): a014241.
  + Hutchinson, L. (2010). "Challenges, controversies, breakthroughs." Nature Reviews Clinical Oncology 7: 669.
  + Koch-Institut, R. (2017). Krebs in Deutschland für 2013/2014, Robert Koch-Institut.
  + Shimoi, T., A. Hamada, M. Yamagishi, M. Hirai, M. Yoshida, T. Nishikawa, K. Sudo, A. Shimomura, E. Noguchi, M. Yunokawa, K. Yonemori, C. Shimizu, T.   + Kinoshita, T. Fukuda, Y. Fujiwara and K. Tamura (2018). "PIK3CA mutation profiling in patients with breast cancer, using a highly sensitive detection system." Cancer science 109(8): 2558-2566.
  + Siraj, A. K., P. Pratheeshkumar, S. K. Parvathareddy, S. P. Divya, F. Al-Dayel, A. Tulbah, D. Ajarim and K. S. Al-Kuraya (2018). "Overexpression of PARP is an independent prognostic marker for poor survival in Middle Eastern breast cancer and its inhibition can be enhanced with embelin co-treatment." Oncotarget 9(99): 37319-37332.
  + Slamon, D., W. Godolphin, L. Jones, J. Holt, S. Wong, D. Keith, W. Levin, S. Stuart, J. Udove, A. Ullrich and a. et (1989). "Studies of the HER-2/neu proto-oncogene in human breast and ovarian cancer." Science 244(4905): 707-712.
  + Stratton, M. R., P. J. Campbell and P. A. Futreal (2009). "The cancer genome." Nature 458(7239): 719-724.
  + Xu, J., Y. Chen and O. I. Olopade (2010). "MYC and Breast Cancer." Genes & cancer 1(6): 629-640.

