---
title: "Linear regression"
author: "Elias Benjamin Farr"
date: "17 Mai 2019"
output: tufte::tufte_handout
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### Multilinear Regression
## Objective

  + For the regression model, we want to predict the CERES score of the driver mutation (dependent variable) based on the CERES score of the SST candidates (independent variables).
  + A multilinear regression will be designed:


# Step 1: Construct datasets for Regression
  + Multilinear regression, we 
    1. reform the BCCL CERES matrix
      + By transponing and format change, we obtain a matrix with colnames = Gene names and rownames = Cell line indices (equal to tPatients_ID)
    2. Collect the relevant the CERES scores for the mutations for the linear regression.
      + All columns with the corresponding names of the SST genes plus the driver mutation will be collected and copied into a new dataframe
      
```{r}
#reformat BCCL CERES matrix
BCCL_kd.ceres_df <- as.data.frame(t(BCCL_kd.ceres))
rownames(BCCL_kd.ceres_df) <- tPatients_ID
colnames(BCCL_kd.ceres_df) <- rownames(BCCL_kd.ceres)

#generate a dataframe containing data for each DV
linRegData_ERBB2 <- BCCL_kd.ceres_df %>% select(rownames(SST_cand_ERBB2), "ERBB2")
linRegData_MYCBP <- BCCL_kd.ceres_df %>% select(rownames(SST_cand_MYCBP), "MYCBP")
linRegData_PARP10 <- BCCL_kd.ceres_df %>% select(rownames(SST_cand_PARP10), "PARP10")
linRegData_PIK3CA <- BCCL_kd.ceres_df %>% select(rownames(SST_cand_PIK3CA), "PIK3CA")

data_regression <- list(linRegData_ERBB2, linRegData_MYCBP, linRegData_PARP10, linRegData_PIK3CA)
head(data_regression)
```

#Step 2: qqplot
  + Check Normality with a qqPlot
    + Using the aes-function of the ggplt package for _aesthetic_ plots.
    
```{r}
create_qqplot <- function(a){  
  p <- ggplot(data_regression[[a]], aes(sample=data_regression[[a]][,ncol(data_regression[[a]])])) + stat_qq()+ stat_qq_line()
   p + labs(title = "QQ-Plot", subtitle = "Check if data is approx. normally distributed")
}
lapply(seq_along(data_regression), create_qqplot)
```

#Step 3: Multilinear Regression: Data split, Model calculation, Return correlation data for evaluation
To run the Multilinear Regression, it is advised to create a long function which
  1. Splits the dataset in 75:25 training:test data
    + Previous installation of the "caTools" package is necessary
  2. Conducts Training of Regression model with training data
    + predict Driver mutation CERES (y; dep.variable) from SST CERES scores (x1 ... xn; indep.variables)
  3. Runs Test Data on model for performance evaluation
  4. Calculates Spearman correlation to further asess model performance
  
```{r}
input_data <- data_regression

lapply(seq_along(input_data), function(a) {
  set.seed(123) #initialize the random numbers, better reproducable
  data <- input_data[[a]] #get the data
  colnames(data)[length(colnames(data))] <- "Predictor"
  
  split = sample.split(data[,ncol(data)], SplitRatio = 0.75) #split the dataset into 3/4 Training and 1/4 Testing dataset
  training_set = subset(data, split == TRUE) #use the labels to get the training data
  test_set = subset(data, split == FALSE)

  regressor = lm(formula = Predictor ~ ., 
                 data = training_set)  #predict CERES of Driver Mutation based on all (=.) the input variables (SSTs)
  
  y_pred = predict(regressor, newdata = test_set) #predict the Driver mut. CERES score based on the test data
  test_set$Prediction = y_pred #adding predictions to the dataset
  corVal <- cor.test(test_set$Predictor , test_set$Prediction, method = "spearman")#to judge model performance, calculate Spearman corr. between predicted and observed values
  return(corVal)
})
```

## Interpretation of Spearman correlation coefficient

# Revision: Meaning of the correlation coefficient rho
  + the rank-correlation coefficient can take values from +1 to -1
    + rho ~ 1: perfect positive association of ranks
    + rho ~ 0: no association of ranks
    + rho ~ -1: perfect negative association of ranks

# Revision: Adding a statistical parameter
  + To further interprete the p-value, we construct a H0 / H1 hypothesis:
    + H0: There is no association between the predicted and observed variable.
    + H1: There is an association between the predicted and observed variable.
  + Most importantly, it is not possible to assess the strength of the correlation based on the p-value. It rather judges the statistical test than the resulting value.
    + as a standard, we set the significance level to 0.05

# Objective
 + We will the compare this values for the different models
  + For a positive correaltion, two variables move in the same direction.
    + For a well-designed model, we expect the precicted and observed CERES scores being approx. even
  + Thus, optimal model design is achieved if:
      + the correlation coefficient is close to 1
      + the p-value < 0.05
    

# Results
We obtained
  + For ERBB2-model:
    + p-value = 0.8397
    + rho: 0.1071
    
  + For MYCBP-model:
    + p-value = 0.7825
    + rho: -0.1428571
    
  + For PARP10-model:
    + p-value = 0.1095
    + rho: -0.6785714
    
  + For PIK3CA-model:
    + p-value = 0.7131
    + rho: -0.1785714

# Discussion

  + ERBB2-model:
    + Shows the highest rho value of all four models, but ~ 0.1 is a rather weak positive relationship
    + The p-value leads to __acceptance of H0__.
    + The data is not even approx. normal distributed.
    
  + MYCBP-model:
    + The rho value leads to very weak negative correlation.
    + A p-value of ~0.78 leads to condifent acceptance of H0.
    + Data is approx. normal distributed.
    
  + PARP10-model:
    + The rho value of ~ -0.67 is a moderate negative relationship.
    + The p-value leads to acceptance of H0. The p-value is the lowest with ~ 0.11. But even if we would raise the confidence level to 0.10 (which would be a questionable move in statistics), the result wouldn´t be significant to accept H1.
    + Data is approx. normal distributed.
    
  + PIK3CA-model:
    + The rho value leads to very weak negative correlation.
    + A p-value of ~0.78  leads to condifent acceptance of H0.

+ __All in all, all multilinear regression models failed to deliver a p-value of significant relevance. Thus, they should not be used for further predictions.__
+ The models for MYCBP and PIK3CA only showed weak, negative correlation, which is not useful for further work with this model.
+ The PARP10 model resulted in a moderate negative relationship, which is not of relevance for our objective.
+ The ERBB2 model did not deliver a significant p-value, but indeed contains at least a weak positive correlation.