---
title: "Linear regression"
author: "Elias Benjamin Farr"
date: "17 Mai 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We want to predict Ceres scores of our drivermutations ("ERBB2","MYCBP", "PARP10", "PIK3CA") based on data we got from second site targets wright ?
for this we need a data frame for every drivermutation, consisting of ceres scores from our SSTs. 
rownames -> SSTs
colnames -> BCCL harvesting this driver mutation ( we can discuss this :))


```{r}
SSTs <- c(rownames(finalFrame)) #could be any list of SSTs
LinResdf0 <- BCCL_kd.ceres[SSTs,] #df, consisting 28 col, nSTTs rows
# something like that: drivmutCL <- c(BCCL_Mutation["Hugosymbol"" == "ERBB2" ])
LinResdf1 <- subset(LinResdf0, colname == drivmutCL)


```
## Idea suggestions for Regression model task

Generally, it was recommended to split the dataset in training:test data (e.g. 80:20) to design a model.

##Training data
Designs linear regression
# Step 1: Quick overview
  + Scatter plot; checks if relation is linear and additive
  + boxplot; detection of strong outliers (I don´t really expect this, but could still be done)
  + density plot; checks if response variable is close to normal distribution (maybe useful for F-test?)
  
# Step 2: Design and linear regression
  + with ln() function; optional insertion of 95% CI
  
# Step 3: evaluate model
  + Is hypothesis supported? (-> estimate, std. error, p-value)
  + Check model fit
    + Residuals should me symm. dist. around 0 (bell shape, use histogram [ggplot2])
    + multiple R squares, should be close to 1 on 0-1 range
    + F-statistics: use p-value to interprete signaficiance for model to fit data
    
If these factors are verified:

## Prediction with test data
  + run model with test data, then calculate prediction accuracy and error rates
  
## Ideas collected from:
[linked phrase](https://www.dataquest.io/blog/statistical-learning-for-predictive-modeling-r/)
[linked phrase](http://r-statistics.co/Linear-Regression.html#Predicting%20Linear%20Models)
