---
title: "Main Code"
author: "Elias Benjamin Farr, Lisa Marie Milchsack"
date: "17 Mai 2019"
output: html_document
---
#installation of required packages
```{r install ggplot2}
# requires installation of latest Rtool version
# install package ggplot2
install.packages("ggplot2")
# load package
library("ggplot2")

#required packages for selection of DV and SSTs
install.packages("tidyverse")
install.packages("factoextra")
install.packages("data.table")
library(tidyverse)
library(cluster)
library(factoextra)
library(data.table)
```

#data clean up
```{r}
#finding all breast cancer cell lines
BCCL_Numbers <- which(allDepMapData$annotation[4]== "Breast Cancer")
BCCL_Annotation <- subset(allDepMapData$annotation, Primary.Disease == "Breast Cancer")
#see what we got
summary(BCCL_Annotation)
# copy expression matrix to new dataframes
df1 <- allDepMapData$expression
df2 <- allDepMapData$copynumber
df3 <- allDepMapData$kd.ceres
df4 <- allDepMapData$kd.prob
#change column names to numbers for easier handling
names(df1) <- c(1:544)
names(df2) <- c(1:544)
names(df3) <- c(1:544)
names(df4) <- c(1:544)
#extract BCCL as dataframe
BCCL_Expression <- as.data.frame( df1[, c(BCCL_Numbers)])
BCCl_Copynumber <- as.data.frame( df2[, c(BCCL_Numbers)])
BCCL_kd.ceres <- as.data.frame( df3[, c(BCCL_Numbers)])
BCCL_kd.prob <- as.data.frame( df4[, c(BCCL_Numbers)])
dim(BCCL_Expression)
dim(BCCl_Copynumber)
dim(BCCL_kd.ceres)
dim(BCCL_kd.prob)
#reorder rows alphabetically
BCCL_Expression <- BCCL_Expression[order(rownames(BCCL_Expression)), ]
BCCl_Copynumber <- BCCl_Copynumber[order(rownames(BCCl_Copynumber)), ]
BCCL_kd.ceres <- BCCL_kd.ceres[order(rownames(BCCL_kd.ceres)), ]
BCCL_kd.prob <- BCCL_kd.prob[order(rownames(BCCL_kd.prob)), ]
#subset Mutation matrix in BCCL
BCCL_Mutation <- allDepMapData$mutation[c(BCCL_Numbers)]
```

```{r} 
Patients_ID <- BCCL_Annotation[1] #get patients ID from annotation matrix
tPatients_ID <- t(Patients_ID) #transpose Patients_ID because it was a column and we want to insert it as row
rownameskdc <- rownames(BCCL_kd.ceres) # save old rownames
existingDF <- BCCL_kd.ceres  # define variables for BCCL_kd.ceres
r <- 1
newrow <- tPatients_ID
#defining a function we can use more often to insert rows
insertRow <- function(existingDF, newrow, r) { 
    existingDF[seq(r+1,nrow(existingDF)+1),] <- existingDF[seq(r,nrow(existingDF)),]
    existingDF[r,] <- newrow
    existingDF
}
BCCL_kd.ceres_ID <- insertRow(existingDF, newrow, r)
rownames(BCCL_kd.ceres_ID) <- c("Deep_ID", rownameskdc) # change rownames back 
BCCL_kd.ceres_ID[1:2, ]
#do the same with other 3 dataframes
rownamesex <- rownames(BCCL_Expression) #save rownames
existingDF <- BCCL_Expression  # define variables for BCCL_Erpression
r <- 1
newrow <- tPatients_ID
BCCL_Expression_ID <- insertRow(existingDF, newrow, r)
rownames(BCCL_Expression_ID) <- c("Deep_ID", rownamesex)
rownamescn <- rownames(BCCl_Copynumber) #save rownames
existingDF <- BCCl_Copynumber  # define variables for BCCL_Copynumber
r <- 1
newrow <- tPatients_ID
BCCl_Copynumber_ID <- insertRow(existingDF, newrow, r)
rownames(BCCl_Copynumber_ID) <- c("Deep_ID", rownamescn)
rownameskdp <- rownames(BCCL_kd.prob) #save rownames
existingDF <- BCCL_kd.prob  # define variables for BCCL_kd.prob
r <- 1
newrow <- tPatients_ID
BCCL_kd.prob_ID <- insertRow(existingDF, newrow, r)
rownames(BCCL_kd.prob_ID) <- c("Deep_ID", rownameskdp)
```

#generation of mutImpact matrix
```{r}
#Generate pre_mutImpact matrix containing all CERES values and with IDs as colnames
pre_mutImpact <- BCCL_kd.ceres
colnames(pre_mutImpact)<-tPatients_ID

#generate empty mutImpact matrix
mutImpact <- matrix( , nrow = 17634, ncol = 28)
colnames(mutImpact)<-colnames(pre_mutImpact)
rownames(mutImpact)<-rownames(pre_mutImpact)

#fill mutImpact matrix with values
for (j in 1:ncol(pre_mutImpact)){
  lineID <- colnames(pre_mutImpact)[j] #select a column-name aka cell line
  for (i in 1:nrow(pre_mutImpact)){
    GOI <- rownames(pre_mutImpact)[i] #select a gene
    if (GOI %in% BCCL_Mutation[[lineID]]$Hugo_Symbol){
      mutImpact[i, j] <- pre_mutImpact[i, j]
    } else {
      mutImpact[i, j] <- NA
    }#replace CERES value with NA in case the gene is not mutated
  }
}
```

#generation of Input Matrix
```{r Input Matrix}
mutImpact_c = mutImpact[rowSums(is.na(mutImpact)) != ncol(mutImpact),]# remove all genes with all NA values
rownamesmutImpc = rownames(mutImpact_c)
mutImpact_c = cbind(rownamesmutImpc, mutImpact_c[,2:ncol(mutImpact_c)]) # insert new column with row names
dim(mutImpact_c) # check gene reduction 
rownameskdc2 = rownames(BCCL_kd.ceres)
BCCL_kd.ceres_2 = cbind(rownameskdc2, BCCL_kd.ceres[,2:ncol(BCCL_kd.ceres)]) # insert new column with row names
BCCL_kd.ceres_3 = BCCL_kd.ceres_2[BCCL_kd.ceres_2[,"rownameskdc2"] %in%  mutImpact_c[,"rownamesmutImpc"],] 
BCCL_kd.ceres_3 = BCCL_kd.ceres_3[,-c(1)]
BCCL_kd.ceres_3 = cbind(BCCL_kd.ceres_3, rowMeans(BCCL_kd.ceres_3)) #calculate rowmeans 
BCCL_kd.ceres_3 = BCCL_kd.ceres_3[rowMeans(BCCL_kd.ceres_3)<= 0,] # deletion of genes with CERES > 0
```

#determining optimal cluster number k
```{r}
BCCL_kd.ceresKS = BCCL_kd.ceres_3[,-c(28)] # remove CERES average column for further processing

CERES_Optimal_K <- function(BCCL_kd.ceresKS, specifier) {
  output <- lapply(1:ncol(BCCL_kd.ceresKS), function(a){
   df <- scale(BCCL_kd.ceresKS[,a]) # pick one column of input data and scale 
   wss <-(nrow(df-1))*sum(apply(df,2,var)) #compute wss
   for (i in 2:15){ # for k between 2 and 15
     set.seed(1234)
     wss[i] <- sum(kmeans(df, centers = i)$withinss)} 
   return(wss)
  })
names (output) <- colnames(BCCL_kd.ceresKS) # rename the output 
return(output)
}

BCCL_kd.ceres_optKS <- CERES_Optimal_K(BCCL_kd.ceresKS, "CERES Optimal Clusters k") 

# plotting wss function 
optimalKprocessedData <- lapply(seq_along(BCCL_kd.ceres_optKS), function(a){
  dtPicker <- as.data.frame(BCCL_kd.ceres_optKS[[a]]) # one vector picked and formatted into dataframe
  dtPicker$Cell_Sample <- names(BCCL_kd.ceres_optKS)[a] #sample added as label
  dtPicker$OptimalK <- 1:nrow(dtPicker) 
  return(dtPicker)
})

optimalKprocessedData <- as.data.frame(rbindlist(optimalKprocessedData)) # output into dataframe
colnames(optimalKprocessedData) <- c("WSS", "Sample", "OptimalK")

# using ggplot to plot output 
ggplot(data = optimalKprocessedData, aes(x=OptimalK, y=WSS))+
  geom_line(aes(color=Cell_Sample))+
  title = ("Optimal number of clusters k") +
  xlab("Number of clusters")+
  ylab("WSS")+
  theme_minimal() +
           theme(legend.position ='bottom',
                  plot.title = element_text(hjust = 0.5),
                  axis.text.x = element_text(angle = 0, vjust = 0, hjust=0.5),
                  legend.title= element_blank(),
                  axis.title.x = element_blank(),
                  strip.text.y = element_text(angle = 0))

# comparison of optimal k means to alternative method  
km_CERES= kmeans(x = BCCL_kd.ceres_3, centers = 2, nstart = 10)
table(km_CERES$cluster)
wss = sapply(2:10, function(k) {
    kmeans(x = BCCL_kd.ceres_3, centers = k)$tot.withinss
})
plot(2:10, wss, type = "b", pch = 19, xlab = "Number of clusters K", ylab = "Total within-clusters sum of squares") 

#comparison of both methods ==> conclusion optimal cluster number k=4

km_CERES= kmeans(x = BCCL_kd.ceres_3, centers = 4, nstart = 10)
table(km_CERES$cluster)
fviz_cluster(km_CERES, data=BCCL_kd.ceres_3, labelsize = 6) # plotting clusters

km_CERES$cluster["ERBB2"]
km_CERES$cluster["ERBB2IP"]
km_CERES$cluster["MYCBP"]
km_CERES$cluster["PARP10"]
km_CERES$cluster["PIK3CA"] 

data_cluster = km_CERES$cluster
rownames(data_cluster) = rownames(BCCL_kd.ceres_3)
data_clus_3 = BCCL_kd.ceres_3[data_cluster == 3,] # without driver mutation label 

drivermut =  c("ERBB2","MYCBP", "PARP10", "PIK3CA") # create vector with driver mutation names
data_clus_3e = data_clus_3
data_clus_3e["Driver mutation"] <- sapply(1:nrow(data_clus_3), function(j){
     if(rownames(data_clus_3e)[j] %in% drivermut){
         data_clus_3e[j,1] = "driver mutation"
     } else {
         data_clus_3e[j,1] = "no driver mutation"
     }
  }) # labeling driver mutations with "driver mutation" in data_clus_3e 
```

# Data visualisation of CERES and TMP values of 4 driver mutations 
```{r}
install.packages("reshape")
install.packages("wesanderson")
library("reshape")
library("wesanderson")

# CERES_boxplot
GOI= BCCL_kd.ceres[c("ERBB2","MYCBP", "PARP10", "PIK3CA"),]
CERES_driver_mut= melt(GOI)
CERES_driver_mut$variable=rep(c("ERBB2","MYCBP", "PARP10", "PIK3CA"), ncol(GOI))

CERES_boxplot = ggplot(CERES_driver_mut, aes(x = variable, y = value)) +
       geom_boxplot(aes(fill=variable), outlier.alpha = 0.7,
                     outlier.colour = "red", outlier.shape = 20, outlier.size = 2) +
        labs(title = 'CERES scores of selected driver mutations', x = 'Selected driver   mutations', y = 'CERES scores') +
            theme_minimal() +
            geom_jitter(width = 0.2) +
  CERES_boxplot = ggplot(CERES_driver_mut, aes(x = variable, y = value)) +
       geom_boxplot(aes(fill=variable), outlier.alpha = 0.7,
                     outlier.colour = "grey", outlier.shape = 20, outlier.size = 2) +
        labs(title = 'CERES scores of selected driver mutations', x = 'Selected driver   mutations', y = 'CERES scores') +
            theme_minimal() +
            geom_jitter(width = 0.2) +
            scale_fill_manual(values=wes_palette(n=4, name="Royal2")) +
            theme(legend.position ='none',
                  plot.title = element_text(hjust = 0.4),
                  axis.text.x = element_text(angle = 0, vjust = 0, hjust=0.5),
                  legend.title= element_blank(),
                  axis.title.x = element_blank(),
                  strip.text.y = element_text(angle = 0))
CERES_boxplot

#Expression_boxplot

GOI_Exp= BCCL_Expression[c("ERBB2","MYCBP", "PARP10", "PIK3CA"),]
Expression_driver_mut= melt(GOI_Exp)
Expression_driver_mut$variable=rep(c("ERBB2","MYCBP", "PARP10", "PIK3CA"), ncol(GOI_Exp))

Expression_boxplot = ggplot(Expression_driver_mut, aes(x = variable, y = value)) +
       geom_boxplot(aes(fill=variable), outlier.alpha = 0.7,
                     outlier.colour = "grey", outlier.shape = 20, outlier.size = 2) +
        labs(title = 'Transcripts per million (TPM) of selected driver mutations', x = 'Selected driver mutations', y = 'Transcripts per million (TPM)') +
            theme_minimal() +
            geom_jitter(width = 0.2) +
            scale_fill_manual(values=wes_palette(n=4, name="Royal2")) +
            theme(legend.position ='none',
                  plot.title = element_text(hjust = 0.4),
                  axis.text.x = element_text(angle = 0, vjust = 0, hjust=0.5),
                  legend.title= element_blank(),
                  axis.title.x = element_blank(),
                  strip.text.y = element_text(angle = 0))
Expression_boxplot
```

#find potential SSTs: Wilcoxon test
```{r}
driver_mut <- c("ERBB2","MYCBP", "PARP10", "PIK3CA") #enter driver mutations
mutImpact_r <- mutImpact[rowSums(is.na(mutImpact)) != ncol(mutImpact), ] #_r=reduced: get rid of all the ONLY NA rows (this will save computation)
mutImpact_kd.ceres <- BCCL_kd.ceres[rownames(BCCL_kd.ceres) %in% rownames(mutImpact_r),] # create matrix containing CERES scores of genes which are mutated once at minimum
colnames(mutImpact_kd.ceres) <- tPatients_ID 

testData <- lapply(seq_along(driver_mut), function(a) {
  driverMutPicker <- driver_mut[a] #pick a driver mutation
  driverMutData <- mutImpact_kd.ceres[driverMutPicker,] #get ceres scores of driver mutation
  
  outputData <- sapply(1:nrow(mutImpact_kd.ceres), function(b) {
    refGeneData <- mutImpact_kd.ceres[b,] #get ceres scores of reference gene (potential SST)
    
    if (rownames(refGeneData) != driverMutPicker) {
      out <- wilcox.test(as.numeric(driverMutData), as.numeric(refGeneData), paired = TRUE)$p.value #get p-value of wilcoxon signed-rank test
      out <- as.data.frame(out, rownames(mutImpact_r)[b]) 
      return(out)
    }
  })
  test = do.call(rbind, outputData)
  rownames(test) <- rownames(mutImpact_r)[which(rownames(mutImpact_r) != driverMutPicker)]
  return(test)
})
names(testData) <- driver_mut

lapply(testData, function(a) head(a)) #look at the data
```

#look for real SSTs based on Wilcoxon test data
```{r}
#get summary of distribution
summary(testData$ERBB2)
summary(testData$MYCBP)
summary(testData$PARP10)
summary(testData$PIK3CA)

#select 15 most promising SST candidates based on highest p-values - libraries needed: dplyr, tibble (included in "tidyverse")

SST_cand_ERBB2 <- as.data.frame(testData$ERBB2, col.names = names(testData$ERBB2)) #create new data frame containing testData for ERBB2
SST_cand_MYCBP <- as.data.frame(testData$MYCBP, col.names = names(testData$MYCBP))
SST_cand_PARP10 <- as.data.frame(testData$PARP10, col.names = names(testData$PARP10))
SST_cand_PIK3CA <- as.data.frame(testData$PIK3CA, col.names = names(testData$PIK3CA))

names(SST_cand_ERBB2) = "ERBB2" #rename column
names(SST_cand_MYCBP) = "MYCBP"
names(SST_cand_PARP10) = "PARP10"
names(SST_cand_PIK3CA) = "PIK3CA"

SST_cand_ERBB2 <- SST_cand_ERBB2 %>%rownames_to_column() %>% top_n(15, ERBB2) %>% column_to_rownames() #select 15 genes with highest p-value
SST_cand_MYCBP <- SST_cand_MYCBP %>% rownames_to_column() %>% top_n(15, MYCBP) %>% column_to_rownames()
SST_cand_PARP10 <- SST_cand_PARP10 %>% rownames_to_column() %>% top_n(15, PARP10) %>% column_to_rownames()
SST_cand_PIK3CA <- SST_cand_PIK3CA %>% rownames_to_column() %>% top_n(15, PIK3CA) %>% column_to_rownames()
```

# Verification of SST candidates by assigned k means cluster number 
```{r}
SST_ERBB2 <- SST_cand_ERBB2 #creating new dataframes
SST_MYCBP <- SST_cand_MYCBP
SST_PARP10 <- SST_cand_PARP10
SST_PIK3CA <- SST_cand_PIK3CA

#labelling of SST candidates with "SST" or "No SST" depending on if in k means cluster 3 or not

#PARP10 candidates
SST_PARP10["SST Cluster Check"] <- sapply(1:nrow(SST_cand_PARP10), function(a){
  if (rownames(SST_cand_PARP10) %in% rownames(data_clus_3e)) {
      SST_PARP10[a,1] ="SST PARP10"
  }else{
     SST_PARP10[a,1] = "No SST"
  }
})
SST_PARP10 <- SST_PARP10 %>% rename(p_values=PARP10)

#ERBB2 candidates
SST_ERBB2["SST Cluster Check"] <- sapply(1:nrow(SST_cand_ERBB2), function(a){
  if (rownames(SST_cand_ERBB2) %in% rownames(data_clus_3e)) {
      SST_ERBB2[a,1] ="SST ERBB2"
  }else{
     SST_PARP10[a,1] = "No SST"
  }
})
SST_ERBB2 <- SST_ERBB2 %>% rename(p_values= ERBB2)

#MYCBP candidates
SST_MYCBP["SST Cluster Check"] <- sapply(1:nrow(SST_cand_MYCBP), function(a){
  if (rownames(SST_cand_MYCBP) %in% rownames(data_clus_3e)) {
      SST_MYCBP[a,1] ="SST MYCBP"
  }else{
     SST_MYCBP[a,1] = "No SST"
  }
})
SST_MYCBP <- SST_MYCBP %>% rename(p_values= MYCBP)

#PIK3CA candidates
SST_PIK3CA["SST Cluster Check"] <- sapply(1:nrow(SST_cand_PIK3CA), function(a){
  if (rownames(SST_cand_PIK3CA) %in% rownames(data_clus_3e)) {
      SST_PIK3CA[a,1] ="SST PIK3CA"
  }else{
     SST_PIK3CA[a,1] = "No SST"
  }
})
SST_PIK3CA <- SST_PIK3CA %>% rename(p_values= PIK3CA)

#All approved SSTs in one dataframe with driver mutation label and p-value
SSTs <-  do.call("rbind", list(SST_ERBB2, SST_MYCBP, SST_PARP10, SST_PIK3CA)) 
SSTs <- SSTs[order(rownames(SSTs)),] # reorder in alphabetical order 

#no SSTs the same for different driver mutations 
```


SST characterization - type of mutation
```{r}
SST_cand_ERBB2$mutType <- sapply(seq_along(rownames(SST_cand_ERBB2)), function(a) {
  SSTPicker <- rownames(SST_cand_ERBB2)[a]
  
  mutTypeData <- c()
  for (i in 1:28) {
    if (SSTPicker %in% BCCL_Mutation[[i]]$Hugo_Symbol) {
      out <- BCCL_Mutation[[i]]$Variant_Classification[a] 
      mutTypeData <- c(mutTypeData, out)
    }
  }
  SST_cand_ERBB2[a, 2] <- paste(mutTypeData, collapse = ", ")
})
```

