---
title: "MutImpactmatrix, k-means, Wilcoxon"
author: "Elias Benjamin Farr"
date: "17 Mai 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Hello lisa :) 
Thank you for this warm welcome, Elias! :)

#requirements to run this code#
in order to run this you will need: BCCL_kd.ceres, tPatients_ID, BCCL_Mutation

#Generate pre_mutImpact matrix containing all ceres values and with IDs as colnames#

```{r} 
pre_mutImpact <- BCCL_kd.ceres
colnames(pre_mutImpact)<-tPatients_ID
```
#Generate loops for generation of mutImpact matrix#

perform exchange for every cell line/ID and select correct list in BCCL_Mutation matrix by checking for colnames of pre_mutImpact matrix:
generate outer loop running from j={1:28} to check for every cell line/ID:
```{r}
colnames(pre_mutImpact)[i]
```

get gene name of gene in row i in pre_mutImpact matrix to look for in BCCL_mutation matrix:
so we can generate an inner loop running from i=(i:17634) looking at each gene 
```{r}
rownames(pre_mutImpact)[i]
```

this might serve as a loop for our purposes:
```{r}
for (j in 1:28){
  lineID <- colnames(pre_mutImpact)[j]
  for (i in 1:17634){
    GOI <- rownames(pre_mutImpact)[i]
    if (GOI %in% BCCL_Mutation[[lineID]]["Hugo_Symbol"]){
      mutImpact[GOI, lineID] <- NA
    } else {
      mutImpact[GOI, lineID] <- pre_mutImpact[GOI, lineID]
    }
  }
}
```

#generate empty mutImpact matrix
```{r}
mutImpact <- matrix( , nrow = 17634, ncol = 28)
colnames(mutImpact)<-colnames(pre_mutImpact)
rownames(mutImpact)<-rownames(pre_mutImpact)
```
