---
title: "MutImpactmatrix, k-means, Wilcoxon"
author: "Elias Benjamin Farr"
date: "17 Mai 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#required packages
```{r}
install.packages("tidyverse")
library(tidyverse)
```

Hello lisa :) 
Thank you for this warm welcome, Elias! :)

#requirements to run this code#
in order to run this you will need: BCCL_kd.ceres, tPatients_ID, BCCL_Mutation



#Generate pre_mutImpact matrix containing all CERES values and with IDs as colnames#
```{r} 
pre_mutImpact <- BCCL_kd.ceres
colnames(pre_mutImpact)<-tPatients_ID
```


#generate empty mutImpact matrix
```{r}
mutImpact <- matrix( , nrow = 17634, ncol = 28)
colnames(mutImpact)<-colnames(pre_mutImpact)
rownames(mutImpact)<-rownames(pre_mutImpact)
```

#fill mutImpact matrix with values
```{r}
for (j in 1:ncol(pre_mutImpact)){
  lineID <- colnames(pre_mutImpact)[j] #select a column-name aka cell line
  for (i in 1:nrow(pre_mutImpact)){
    GOI <- rownames(pre_mutImpact)[i] #select a gene
    if (GOI %in% BCCL_Mutation[[lineID]]$Hugo_Symbol){
      mutImpact[i, j] <- pre_mutImpact[i, j]
    } else {
      mutImpact[i, j] <- NA
    }#replace CERES value with NA in case the gene is not mutated
  }
}
```

##Paired Wilcoxon signed rank test

```{r}
#generate matrix of driver mutations
drivMat <- mutImpact[ , c(#colnr of DV)]

#generate 2nd vector of potential SSTS, i = gene
potSST <- as.vector(mutImpact[a, ])

#select driver mutation to find correspaonding SSTs
for i in (1:ncol(drivMat) function() {
  Dvm <- as.vector(drivMat[ ,i]
})

#suggested loop to design Wilcoxon matrix
wlxmat <- lapply(1:nrow(mutImpact), function(a) {
  potSST <- as.vector(mutImpact[a, ])
  wilcox.test(Dvm, potSST, paired = TRUE, na.action = na.exclude)
})
#returns error:  Error in wilcox.test.default(Dvm, potSST, paired = TRUE, na.action = na.exclude) : 
  #not enough (finite) 'x' observations 

#new try... "by foot"
wlxmat <- sapply(1:nrow(mutImpact), function(a) {
  wil.p = 0
  wil.p.sum = 0
  non.na = 0
  for (i in 1:ncol(mutImpact)) {
    if (!is.na(mutImpact["BPI",i]) & !is.na(mutImpact[a,i]) & mutImpact["BPI",i]!=mutImpact[a,i]) {#here BPI is supposed to be a DV
        wil.p = wilcox.test(mutImpact["BPI",i],mutImpact[a,i], paired = TRUE)$p.value
        non.na = non.na + 1
    } else {
        wil.p = 0
    }
    wil.p.sum = wil.p.sum + wil.p
}
wil.p.mean = wil.p.sum/non.na
})

#problem is that you cannot change rownames and there is something wrong with the computation as some values are 1 (NAN are the values in case there is no co-existence of the DV and the SST)
```
##unpaired Wilcoxon test:
```{r}
driver_mut <- c("ERBB2","MYCBP", "PARP10", "PIK3CA") #enter driver mutations
mutImpact_r = mutImpact[rowSums(is.na(mutImpact)) != ncol(mutImpact), ] #_r=reduced: get rid of all the ONLY NA rows (this will save computation)

finalData <- lapply(seq_along(driver_mut), function(a) {
  driverMutPicker <- driver_mut[a] #pick a driver mutation
  outputData <- sapply(1:nrow(mutImpact_r), function(b) {
    row_picker <- mutImpact_r[b,]
    
    #row_picker <- row_picker[complete.cases(row_picker)] #pick refernce data and get rid of NA values
    #print(paste0("I am doing: ",a, ";", b))
    #Extract the reference
    refData <- mutImpact_r[which(rownames(mutImpact_r) == driverMutPicker),] #pick the data for the reference driver Mutation
    #refData <- refData[complete.cases(refData)] #make sure to take no NA values
    
    #Make a wilcoxon test --> make an unpaired test: https://stats.stackexchange.com/questions/93734/paired-or-unpaired-wilcoxon-test
    out <- wilcox.test(as.vector(refData[complete.cases(refData)]), as.vector(row_picker[complete.cases(row_picker)]))
    #out <- wilcox.test(refData, row_picker, paired = TRUE, na.action = na.exclude)
    return(out$p.value)
  })
  names(outputData) <- rownames(mutImpact_r)
  return(outputData)
})
names(finalData) <- driver_mut

#Reformat the data
finalData <- lapply(seq_along(finalData), function(a) {
  out <- as.data.frame(finalData[[a]])
  return(out)
})
#Use the RDS_lists to create the matrix
rownames2col <- function(inDF, RowName = ".rownames") {
  temp <- data.frame(rownames(inDF), inDF, row.names = NULL)
  names(temp)[1] <- RowName
  temp
}

#Put all the sublists to a nice dataframe
finalFrame <- Reduce(function(finalData, y) merge(finalData, y, by = ".rownames", all = TRUE), lapply(finalData, rownames2col))
rownames(finalFrame) <- finalFrame[,1]
finalFrame <- finalFrame[,2:ncol(finalFrame)]
colnames(finalFrame) <- driver_mut
head(finalFrame)
```
#get an overview over SST status
```{r}
#generate matrix containing all non-SSTs
non_potSSTs <- BCCL_kd.ceres[-which(rownames(BCCL_kd.ceres) == rownames(mutImpact_r)),]

#add information on SST status in BCCL_kd.ceres matrix
BCCL_kd.ceres_an$SST_status <- sapply(1:nrow(BCCL_kd.ceres_an), function(c) {
  if (rownames(BCCL_kd.ceres_an)[c] %in% rownames(mutImpact_r)) {
    BCCL_kd.ceres_an[c, 2] = "pot SST"
  } else {
    BCCL_kd.ceres_an[c, 2] = "no SST"
  }
})
BCCL_kd.ceres_an <- BCCL_kd.ceres_an[c("mutation", "SST_status", tPatients_ID)]#reorder columns
```
#look for real SSTs based on Wilcoxon test data
```{r}
#visualize distribution of p-vlaues for each DV
par(mfrow =c(1,4))
plot(finalFrame$ERBB2, type = "p", pch = 19, xlab = "gene Index", ylab = "mean p-value Wilcoxon Test" )
plot(finalFrame$MYCBP, type = "p", pch = 19, xlab = "gene Index", ylab = "mean p-value Wilcoxon Test" )
plot(finalFrame$PARP10, type = "p", pch = 19, xlab = "gene Index", ylab = "mean p-value Wilcoxon Test" )
plot(finalFrame$PIK3CA, type = "p", pch = 19, xlab = "gene Index", ylab = "mean p-value Wilcoxon Test" )

#get summary of distribution
summary(finalFrame$ERBB2)
summary(finalFrame$MYCBP)
summary(finalFrame$PARP10)
summary(finalFrame$PIK3CA)

#select eight most promising SST candidates based on highest p-values - libraries needed: dplyr, tibble (included in "tidyverse") -> I will try and find a simpler version for this code.. I am pretty sure you can handle this with an apply command
SST_cand_ERBB2 <- finalFrame %>% select("ERBB2") %>% rownames_to_column() %>% filter(ERBB2 > quantile(ERBB2, probs = 0.999)) %>% column_to_rownames()
SST_cand_MYCBP <- finalFrame %>% select("MYCBP") %>% rownames_to_column() %>% filter(MYCBP > quantile(MYCBP, probs = 0.999)) %>% column_to_rownames()
SST_cand_PARP10 <- finalFrame %>% select("PARP10") %>% rownames_to_column() %>% filter(!is.na(PARP10)) %>% filter(PARP10 > quantile(PARP10, probs = 0.999)) %>% column_to_rownames() #removal of NA required 
SST_cand_PIK3CA <- finalFrame %>% select("PIK3CA") %>% rownames_to_column() %>% filter(PIK3CA > quantile(PIK3CA, probs = 0.999)) %>% column_to_rownames()
```


=======
```{r include=FALSE}
table(is.na(mutImpact)) #counting the NAs over the whole mutImpact matrix.
PCA_mutimpact <- mutImpact #creating a copy
dim(PCA_mutimpact) #removing all NAs by a for loop and replacing them with a zero
for (i in 1:17634){ #for all genes (rows)
  for(j in 1:28){ #for all cell lines (columns)
    if (is.na(PCA_mutimpact[i,j])){
      PCA_mutimpact[i,j] <- 0
    }
  }
}
anyNA(PCA_mutimpact) #verify if all NAs are removed
scale(PCA_mutimpact, center = TRUE, scale = TRUE) #unit variance as PCA preperation
mutImpact.pca = prcomp(PCA_mutimpact, center = F, scale. = F) #set the parameters false since we already centered and scaled the data
print(mutImpact.pca)
plot(mutImpact.pca, type = "l") #scree plot to detect the PCs explaining most of the variance: PC 1 and 2 have each 3-4% explained variances, PC3-8 approx. 1% each. We ttry to continue with PC 1/2.
install.packages("ggfortify")
library(ggfortify)
df <- pca[c(1,2)]
df <- as.data.frame(sapply(pca[c(1,2)], scale))#to eliminate the error "Error in colMeans(x, na.rm = TRUE) : 'x' must be numeric", since cateogorical variables cannot be sclaed
autoplot(prcomp(df), loadings = TRUE, loadings.colour = "turquoise", loadings.label = TRUE, loadings.label.size = 3) #we receive the PCA plot
```

#new try - identification of SSTs
```{r}
driver_mut <- c("ERBB2","MYCBP", "PARP10", "PIK3CA") #enter driver mutations
mutImpact_r = mutImpact[rowSums(is.na(mutImpact)) != ncol(mutImpact), ] #_r=reduced: get rid of all the ONLY NA rows (this will save computation)
mutImpact_df <- as.data.frame(t(mutImpact_r)) #generate data frame based on reduced mutImpact matrix

finalData <- lapply(seq_along(driver_mut), function(a) {
  driverMutPicker <- driver_mut[a] #pick a driver mutation
  outputData <- sapply(1:ncol(mutImpact_df), function(b) {
    refgene_picker <- mutImpact_df[,b] #pick reference gene
    refData <- mutImpact_df %>% select(driverMutPicker, b) #select values of driver mutation and reference gene
    
    #check whether mutations co-occur
    if (rowSums(is.na(refData)) == 0) {
      refData_subset <- refData %>% drop.na()
      out <- wilcox.test(as.vector(refData_subset[,1]), as.vector(refData_subset[,2]), paired = TRUE)
    } else {
      out <- NA
    }
    return(out)
  })
  names(outputData) <- rownames(mutImpact_r)
  return(outputData)
})
names(finalData) <- driver_mut

#error: the condition has length > 1 and only the first element will be usedError: $ operator is invalid for atomic vectors
```

#new try based on reduced CERES madtrix - identification of SSTs
```{r}
driver_mut <- c("ERBB2","MYCBP", "PARP10", "PIK3CA") #enter driver mutations
mutImpact_r <- mutImpact[rowSums(is.na(mutImpact)) != ncol(mutImpact), ] #_r=reduced: get rid of all the ONLY NA rows (this will save computation)
mutImpact_kd.ceres <- BCCL_kd.ceres[rownames(BCCL_kd.ceres) %in% rownames(mutImpact_r),] # create matrix containing CERES scores of genes which are mutated once at minimum
colnames(mutImpact_kd.ceres) <- tPatients_ID 

testData <- lapply(seq_along(driver_mut), function(a) {
  driverMutPicker <- driver_mut[a] #pick a driver mutation
  driverMutData <- mutImpact_kd.ceres[driverMutPicker,] #get ceres scores of driver mutation
  
  outputData <- sapply(1:nrow(mutImpact_kd.ceres), function(b) {
    refGeneData <- mutImpact_kd.ceres[b,] #get ceres scores of reference gene (potential SST)
    
    if (rownames(refGeneData) != driverMutPicker) {
      out <- wilcox.test(as.numeric(driverMutData), as.numeric(refGeneData), paired = TRUE)$p.value #get p-value of wilcoxon signed-rank test
      return(out)
    }
  })
  names(outputData) <- rownames(mutImpact_r)
  return(outputData)
})
names(testData) <- driver_mut

#Reformat the data
testData <- lapply(seq_along(testData), function(a) {
  out <- as.data.frame(testData[[a]])
  return(out)
})

#Use the RDS_lists to create the matrix
rownames2col <- function(inDF, RowName = ".rownames") {
  temp <- data.frame(rownames(inDF), inDF, row.names = NULL)
  names(temp)[1] <- RowName
  temp
}

#Put all the sublists to a nice dataframe
testFrame <- Reduce(function(testlData, y) merge(testData, y, by = ".rownames", all = TRUE), lapply(testData, rownames2col))
rownames(testFrame) <- testFrame[,1]
testFrame <- testFrame[,2:ncol(testFrame)]
colnames(testFrame) <- driver_mut
head(testFrame)
```

## final version
#2nd new try based on reduced CERES madtrix - identification of SSTs
```{r}
driver_mut <- c("ERBB2","MYCBP", "PARP10", "PIK3CA") #enter driver mutations
mutImpact_r <- mutImpact[rowSums(is.na(mutImpact)) != ncol(mutImpact), ] #_r=reduced: get rid of all the ONLY NA rows (this will save computation)
mutImpact_kd.ceres <- BCCL_kd.ceres[rownames(BCCL_kd.ceres) %in% rownames(mutImpact_r),] # create matrix containing CERES scores of genes which are mutated once at minimum
colnames(mutImpact_kd.ceres) <- tPatients_ID 

testData <- lapply(seq_along(driver_mut), function(a) {
  driverMutPicker <- driver_mut[a] #pick a driver mutation
  driverMutData <- mutImpact_kd.ceres[driverMutPicker,] #get ceres scores of driver mutation
  
  outputData <- sapply(1:nrow(mutImpact_kd.ceres), function(b) {
    refGeneData <- mutImpact_kd.ceres[b,] #get ceres scores of reference gene (potential SST)
    
    if (rownames(refGeneData) != driverMutPicker) {
      out <- wilcox.test(as.numeric(driverMutData), as.numeric(refGeneData), paired = TRUE)$p.value #get p-value of wilcoxon signed-rank test
      out <- as.data.frame(out, rownames(mutImpact_r)[b]) 
      return(out)
    }
  })
  test = do.call(rbind, outputData)
  rownames(test) <- rownames(mutImpact_r)[which(rownames(mutImpact_r) != driverMutPicker)]
  return(test)
})
names(testData) <- driver_mut

lapply(testData, function(a) head(a)) #look at the data


```
#look for real SSTs based on Wilcoxon test data
```{r}
#get summary of distribution
summary(testData$ERBB2)
summary(testData$MYCBP)
summary(testData$PARP10)
summary(testData$PIK3CA)

#select 15 most promising SST candidates based on highest p-values - libraries needed: dplyr, tibble (included in "tidyverse")

SST_cand_ERBB2 <- as.data.frame(testData$ERBB2, col.names = names(testData$ERBB2)) #create new data frame containing testData for ERBB2
SST_cand_MYCBP <- as.data.frame(testData$MYCBP, col.names = names(testData$MYCBP))
SST_cand_PARP10 <- as.data.frame(testData$PARP10, col.names = names(testData$PARP10))
SST_cand_PIK3CA <- as.data.frame(testData$PIK3CA, col.names = names(testData$PIK3CA))

names(SST_cand_ERBB2) = "ERBB2" #rename column
names(SST_cand_MYCBP) = "MYCBP"
names(SST_cand_PARP10) = "PARP10"
names(SST_cand_PIK3CA) = "PIK3CA"

SST_cand_ERBB2 <- SST_cand_ERBB2 %>%rownames_to_column() %>% top_n(15, ERBB2) %>% column_to_rownames() #select 15 genes with highest p-value
SST_cand_MYCBP <- SST_cand_MYCBP %>% rownames_to_column() %>% top_n(15, MYCBP) %>% column_to_rownames()
SST_cand_PARP10 <- SST_cand_PARP10 %>% rownames_to_column() %>% top_n(15, PARP10) %>% column_to_rownames()
SST_cand_PIK3CA <- SST_cand_PIK3CA %>% rownames_to_column() %>% top_n(15, PIK3CA) %>% column_to_rownames()
```

#shorter version: SST selection
```{r}
SST_Data <- lapply(seq_along(testData), function(a) {
  data <- as.data.frame(testData[[a]], col.names = names(testData[[a]])) #create new data frame containing testData for each SST
  colnames(data) = driver_mut[a] #rename columns
  data <- data %>%rownames_to_column() %>% top_n(15, data[,1]) %>% column_to_rownames() #select 15 genes with highest p-value
  return(data)
})
names(SST_Data) <- driver_mut
lapply(SST_Data, function(a) print(a)) #have a look at the SSTs
```

#SST characterization - type of mutation
```{r}
SST_cand_ERBB2$mutType <- sapply(seq_along(rownames(SST_cand_ERBB2)), function(a) {
  SSTPicker <- rownames(SST_cand_ERBB2)[a]
  
  mutTypeData <- c()
  for (i in 1:28) {
    if (SSTPicker %in% BCCL_Mutation[[i]]$Hugo_Symbol) {
      out <- BCCL_Mutation[[i]]$Variant_Classification[a] 
      mutTypeData <- c(mutTypeData, out)
    }
  }
  SST_cand_ERBB2[a, 2] <- paste(mutTypeData, collapse = ", ")
})
```

#SST characterization - type of mutation (automated)
```{r}
SST_Data <- lapply(seq_along(SST_Data), function(a) {
  dvPicker <- SST_Data[[a]]
  dvPicker$mutType <- sapply(seq_along(rownames(dvPicker)), function(b) {
    SSTPicker <- rownames(dvPicker)[b]
    mutTypeData <- c()
    for (i in 1:ncol(BCCL_kd.ceres)) { 
      if (SSTPicker %in% BCCL_Mutation[[i]]$Hugo_Symbol) {
        out <- BCCL_Mutation[[i]]$Variant_Classification[a] 
        mutTypeData <- c(mutTypeData, out)
      }
    }
    dvPicker[b, 2] <- paste(mutTypeData, collapse = ", ")
  })
  return(dvPicker)
})
names(SST_Data) <- driver_mut
```