---
title: "kmeans -> Dmut"
author: "Salome Steinke"
date: "17 Mai 2019"
output:
  pdf_document: default
  html_document: default
---

#Loading Data
```{r Load.Data}
allDepMapData = readRDS("C:\Users\Salome\Documents\4.FS\Bioinfo")
BCCL_Numbers = which(allDepMapData$annotation[4]== "Breast Cancer") 
BCCL_Annotation = subset(allDepMapData$annotation, Primary.Disease == "Breast Cancer")
#see what we got
summary(BCCL_Annotation)
# copy expression matrix to new dataframes
df1 <- allDepMapData$expression
df2 <- allDepMapData$copynumber
df3 <- allDepMapData$kd.ceres
df4 <- allDepMapData$kd.prob
#change column names to numbers for easier handling
names(df1) <- c(1:544)
names(df2) <- c(1:544)
names(df3) <- c(1:544)
names(df4) <- c(1:544)
#extract BCCL as dataframe
BCCL_Expression <- as.data.frame( df1[, c(BCCL_Numbers)])
BCCl_Copynumber <- as.data.frame( df2[, c(BCCL_Numbers)])
BCCL_kd.ceres <- as.data.frame( df3[, c(BCCL_Numbers)])
BCCL_kd.prob <- as.data.frame( df4[, c(BCCL_Numbers)])
dim(BCCL_Expression)
dim(BCCl_Copynumber)
dim(BCCL_kd.ceres)
dim(BCCL_kd.prob)
#reorder rows alphabetically
BCCL_Expression <- BCCL_Expression[order(rownames(BCCL_Expression)), ]
BCCl_Copynumber <- BCCl_Copynumber[order(rownames(BCCl_Copynumber)), ]
BCCL_kd.ceres <- BCCL_kd.ceres[order(rownames(BCCL_kd.ceres)), ]
BCCL_kd.prob <- BCCL_kd.prob[order(rownames(BCCL_kd.prob)), ]
#subset Mutation matrix in BCCL
BCCL_Mutation <- allDepMapData$mutation[c(BCCL_Numbers)]

Patients_ID <- BCCL_Annotation[1] #get patients ID from annotation matrix
tPatients_ID <- t(Patients_ID) #transpose Patients_ID because it was a column and we want to insert it as row
rownameskdc <- rownames(BCCL_kd.ceres) # save old rownames
existingDF <- BCCL_kd.ceres  # define variables for BCCL_kd.ceres
r <- 1
newrow <- tPatients_ID
#defining a function we can use more often to insert rows
insertRow <- function(existingDF, newrow, r) { 
    existingDF[seq(r+1,nrow(existingDF)+1),] <- existingDF[seq(r,nrow(existingDF)),]
    existingDF[r,] <- newrow
    existingDF
}
BCCL_kd.ceres_ID <- insertRow(existingDF, newrow, r)
rownames(BCCL_kd.ceres_ID) <- c("Deep_ID", rownameskdc) # change rownames back 
BCCL_kd.ceres_ID[1:2, ]
#do the same with other 3 dataframes
rownamesex <- rownames(BCCL_Expression) #save rownames
existingDF <- BCCL_Expression  # define variables for BCCL_Erpression
r <- 1
newrow <- tPatients_ID
BCCL_Expression_ID <- insertRow(existingDF, newrow, r)
rownames(BCCL_Expression_ID) <- c("Deep_ID", rownamesex)
rownamescn <- rownames(BCCl_Copynumber) #save rownames
existingDF <- BCCl_Copynumber  # define variables for BCCL_Copynumber
r <- 1
newrow <- tPatients_ID
BCCl_Copynumber_ID <- insertRow(existingDF, newrow, r)
rownames(BCCl_Copynumber_ID) <- c("Deep_ID", rownamescn)
rownameskdp <- rownames(BCCL_kd.prob) #save rownames
existingDF <- BCCL_kd.prob  # define variables for BCCL_kd.prob
r <- 1
newrow <- tPatients_ID
BCCL_kd.prob_ID <- insertRow(existingDF, newrow, r)
rownames(BCCL_kd.prob_ID) <- c("Deep_ID", rownameskdp)
```

# Identification of driver mutations using k means clustering of CERES scores 

In order to verify our chosen driver mutations from literature, k means clustering was performed.


```{r kmeans ceres}
#loading required packages
install.packages("tidyverse")
install.packages("factoextra")
library(tidyverse)
library(cluster)
library(factoextra)

#Loading mutImpact matrix
pre_mutImpact <- BCCL_kd.ceres
colnames(pre_mutImpact)<-tPatients_ID
mutImpact <- matrix( , nrow = 17634, ncol = 28)
colnames(mutImpact)<-colnames(pre_mutImpact)
rownames(mutImpact)<-rownames(pre_mutImpact)
for (j in 1:ncol(pre_mutImpact)){
  lineID <- colnames(pre_mutImpact)[j] #select a column-name aka cell line
  for (i in 1:nrow(pre_mutImpact)){
    GOI <- rownames(pre_mutImpact)[i] #select a gene
    if (GOI %in% BCCL_Mutation[[lineID]]$Hugo_Symbol){
      mutImpact[i, j] <- pre_mutImpact[i, j]
    } else {
      mutImpact[i, j] <- NA
    }#replace CERES value with NA in case the gene is not mutated
  }
}
````

K means clustering was performed with all genes mutated at least once out of all 28 cell line samples and whose CERES score is equal to or lower than zero. For this, all genes with only NA values, hence not one mutation in any cell sample, were removed. A new dataframe `BCCL_kd.ceres_2`was generated containing all 28 CERES scores for each gene, with gene names in the first column. Gene names in the first column of the data frame `BCCL_kd.ceres_2`were compared to the rownames of `mutImpact_c` to generate the BCCL_kd.ceres_3 dataframe containing CERES scores for all breast cell cancer lines for all genes mutated at least once. Subsequently, the mean CERES score for each gene was computed using the rowmeans function and saved in the last column. For this, the column `rownameskdc2`had to be removed. Finally, all genes with CERES scores > 0 were deleted. The final BCCL_kd.ceres_3 dataframe now contains 4934 genes mutated at least once in the 28 breast cancer cell lines and with CERES scores equal to or below 0. 

```{r}
# selecting only genes mutated at least once and with Ceres >= 0
mutImpact_c = mutImpact[rowSums(is.na(mutImpact)) != ncol(mutImpact),]# remove all genes withall NA values
rownamesmutImpc = rownames(mutImpact_c)
mutImpact_c = cbind(rownamesmutImpc, mutImpact_c[,2:ncol(mutImpact_c)]) # insert new column with row names
dim(mutImpact_c) # check gene reduction 
rownameskdc2 = rownames(BCCL_kd.ceres)
BCCL_kd.ceres_2 = cbind(rownameskdc2, BCCL_kd.ceres[,2:ncol(BCCL_kd.ceres)]) # insert new column with row names
BCCL_kd.ceres_3 = BCCL_kd.ceres_2[BCCL_kd.ceres_2[,"rownameskdc2"] %in%  mutImpact_c[,"rownamesmutImpc"],] 
BCCL_kd.ceres_3 = BCCL_kd.ceres_3[,-c(1)]
BCCL_kd.ceres_3 = cbind(BCCL_kd.ceres_3, rowMeans(BCCL_kd.ceres_3)) #calculate rowmeans 
BCCL_kd.ceres_3 = BCCL_kd.ceres_3[rowMeans(BCCL_kd.ceres_3)<= 0,] # deletion of genes with CERES > 0
```

Before conducting k means clustering the column containing average CERES scores for each gene was removed (new dataframe `BCCL_kd.ceresKS`) and the optimal number of k clusters identified. For each column the data values were scaled  

```{r}
#determining optimal cluster number k
BCCL_kd.ceresKS = BCCL_kd.ceres_3[,-c(28)] # remove CERES average column for further processing

CERES_Optimal_K <- function(BCCL_kd.ceresKS, specifier) {
  output <- lapply(1:ncol(BCCL_kd.ceresKS), function(a){
   df <- scale(BCCL_kd.ceresKS[,a]) # pick one column of input data and scale 
   wss <-(nrow(df-1))*sum(apply(df,2,var)) #compute wss
   for (i in 2:15){ # for k between 2 and 15
     set.seed(1234)
     wss[i] <- sum(kmeans(df, centers = i)$withinss)} 
   return(wss)
  })
names (output) <- colnames(BCCL_kd.ceresKS) # rename the output 
return(output)
}

BCCL_kd.ceres_optKS <- CERES_Optimal_K(BCCL_kd.ceresKS, "CERES Optimal Clusters k") 

# plotting wss function 
optimalKprocessedData <- lapply(seq_along(BCCL_kd.ceres_optKS), function(a){
  dtPicker <- as.data.frame(BCCL_kd.ceres_optKS[[a]]) # one vector picked and formatted into dataframe
  dtPicker$Cell_Sample <- names(BCCL_kd.ceres_optKS)[a] #sample added as label
  dtPicker$OptimalK <- 1:nrow(dtPicker) 
  return(dtPicker)
})

optimalKprocessedData <- as.data.frame(rbindlist(optimalKprocessedData)) # output into dataframe
colnames(optimalKprocessedData) <- c("WSS", "Sample", "OptimalK")

# using ggplot to plot output 
ggplot(data = optimalKprocessedData, aes(x=OptimalK, y=WSS))+
  geom_line(aes(color=Cell_Sample))+
  title = ("Optimal number of clusters k") +
  xlab("Number of clusters")+
  ylab("WSS")+
  theme_minimal() +
           theme(legend.position ='bottom',
                  plot.title = element_text(hjust = 0.5),
                  axis.text.x = element_text(angle = 0, vjust = 0, hjust=0.5),
                  legend.title= element_blank(),
                  axis.title.x = element_blank(),
                  strip.text.y = element_text(angle = 0))
```

*Insert Text about comparison to alternative Wss* 
```{r}
# comparison of optimal k means to alternative method  

km_CERES= kmeans(x = BCCL_kd.ceres_3, centers = 2, nstart = 10)
table(km_CERES$cluster)
wss = sapply(2:10, function(k) {
    kmeans(x = BCCL_kd.ceres_3, centers = k)$tot.withinss
})
plot(2:10, wss, type = "b", pch = 19, xlab = "Number of clusters K", ylab = "Total within-clusters sum of squares") 

#comparison of both methods ==> conclusion optimal cluster number k=4 
```

Having identified the optimal cluster number k to be 4, k means clustering of the data in BCCL_kd.ceres_3 was performed with 10 iterations. The following cluster plot was generated *(Insert cluster plot)*.

```{r}
km_CERES= kmeans(x = BCCL_kd.ceres_3, centers = 4, nstart = 10)
table(km_CERES$cluster)
fviz_cluster(km_CERES, data=BCCL_kd.ceres_3, labelsize = 6) # plotting clusters
```

In order to verify our selection of driver mutations from literature, the cluster number of each driver mutation was identified. All selected driver mutations from literature,"ERBB2", "ERBB2IP", "MYCBP", "PARP10", "PIK3CA" and "CCND" were localized in cluster 3, hence with similar CERES scores, with the exception of "CCND" and "ERBBIP". While "CCND" was removed in the filtering process and no longer found in`BCCL_kd.ceres_3`, "ERBB2IP" was found in cluster 2. These two mutations were thus excluded from further analysis. 

Therefore selected driver mutations used in further analysis were "ERBB2", "MYCBP", "PARP10" and "PIK3CA". 

``` {r}
km_CERES$cluster["ERBB2"]
km_CERES$cluster["ERBB2IP"]
km_CERES$cluster["MYCBP"]
km_CERES$cluster["PARP10"]
km_CERES$cluster["PIK3CA"] 
```

Since our approach defined SSTs as genes expected to interact synergistcally with the corresponding driver mutation in promoting cell viability, similar CERES scores were expected. Thus all potential SSTs candidates were expected to be found also in the same cluster as selected driver mutations, hence in cluster 3. A dataframe `data_clus_3`was subsequently generated to contain all 1148 genes colocalized with driver mutations in cluster 3. 

```{r}
data_cluster = km_CERES$cluster
rownames(data_cluster) = rownames(BCCL_kd.ceres_3)
data_clus_3 = BCCL_kd.ceres_3[data_cluster == 3,] #dataframe with all 1148 genes in same cluster as driver mutations 
```

A further dataframe `data_clus_3e` was subsequently generated labelling the four driver mutations in a separate column, by implementing an "IF" loop. If the rowname of `data_clus_3e`was found in a previously created vector containing driver mutations names, it was labelled as "driver mutation".Otherwise genes were labelled as "no driver mutation".

```{r}
#LABELING DRIVER MUTATIONS 
drivermut =  c("ERBB2","MYCBP", "PARP10", "PIK3CA") # create vector with driver mutation names
data_clus_3e = data_clus_3
data_clus_3e["Driver mutation"] <- sapply(1:nrow(data_clus_3), function(j){
     if(rownames(data_clus_3e)[j] %in% drivermut){
         data_clus_3e[j,1] = "driver mutation"
     } else {
         data_clus_3e[j,1] = "no driver mutation"
     }
  }) # labeling driver mutations with "driver mutation" in data_clus_3e 
```

# Data visualisation of CERES and TPM values of the 4 selected driver mutations 

CERES scores and expression as TPM values for the selected 4 driver mutations were visualised as box and whisker plots. Following the generation of the `GOI_X` dataframe containing either CERES or TPM values for the four mutations, the data was reshaped to contain two columns. One "variable" with the names of driver mutations, the other "value" with the corresponding CERES or TPM value. The ggplot function, part of the ggplot2 package, was used to plot the following box an whisker plots. 

```{data visualisation}
install.packages("reshape")
install.packages("wesanderson")
library("reshape")
library("wesanderson")

# CERES_boxplot
GOI= BCCL_kd.ceres[c("ERBB2","MYCBP", "PARP10", "PIK3CA"),]
CERES_driver_mut= melt(GOI)
CERES_driver_mut$variable=rep(c("ERBB2","MYCBP", "PARP10", "PIK3CA"), ncol(GOI))

CERES_boxplot = ggplot(CERES_driver_mut, aes(x = variable, y = value)) +
       geom_boxplot(aes(fill=variable), outlier.alpha = 0.7,
                     outlier.colour = "red", outlier.shape = 20, outlier.size = 2) +
        labs(title = 'CERES scores of selected driver mutations', x = 'Selected driver   mutations', y = 'CERES scores') +
            theme_minimal() +
            geom_jitter(width = 0.2) +
  CERES_boxplot = ggplot(CERES_driver_mut, aes(x = variable, y = value)) +
       geom_boxplot(aes(fill=variable), outlier.alpha = 0.7,
                     outlier.colour = "grey", outlier.shape = 20, outlier.size = 2) +
        labs(title = 'CERES scores of selected driver mutations', x = 'Selected driver   mutations', y = 'CERES scores') +
            theme_minimal() +
            geom_jitter(width = 0.2) +
            scale_fill_manual(values=wes_palette(n=4, name="Royal2")) +
            theme(legend.position ='none',
                  plot.title = element_text(hjust = 0.4),
                  axis.text.x = element_text(angle = 0, vjust = 0, hjust=0.5),
                  legend.title= element_blank(),
                  axis.title.x = element_blank(),
                  strip.text.y = element_text(angle = 0))
CERES_boxplot

  
#Expression_boxplot
GOI_Exp= BCCL_Expression[c("ERBB2","MYCBP", "PARP10", "PIK3CA"),]
Expression_driver_mut= melt(GOI_Exp)
Expression_driver_mut$variable=rep(c("ERBB2","MYCBP", "PARP10", "PIK3CA"), ncol(GOI_Exp))

Expression_boxplot = ggplot(Expression_driver_mut, aes(x = variable, y = value)) +
       geom_boxplot(aes(fill=variable), outlier.alpha = 0.7,
                     outlier.colour = "grey", outlier.shape = 20, outlier.size = 2) +
        labs(title = 'Transcripts per million (TPM) of selected driver mutations', x = 'Selected driver mutations', y = 'Transcripts per million (TPM)') +
            theme_minimal() +
            geom_jitter(width = 0.2) +
            scale_fill_manual(values=wes_palette(n=4, name="Royal2")) +
            theme(legend.position ='none',
                  plot.title = element_text(hjust = 0.4),
                  axis.text.x = element_text(angle = 0, vjust = 0, hjust=0.5),
                  legend.title= element_blank(),
                  axis.title.x = element_blank(),
                  strip.text.y = element_text(angle = 0))
Expression_boxplot
```

*Analysis of box and whisker plots required*