---
title: "PCA and related plots -> Dmut"
author: "Elias Benjamin Farr"
date: "17 Mai 2019"
output: html_document
---
Objective of PCA:
  + For this dimension reduction process, we try to find similarity in the dataset.
  + Prinicipal Components describe the directions with the most varianc in data.
  
-Loading Data
```{r Load.Data}
allDepMapData = readRDS("path/to/your/directory/DepMap19Q1_allData.RDS")
BCCL_Numbers = which(allDepMapData$annotation[4]== "Breast Cancer") 
BCCL_Annotation = subset(allDepMapData$annotation, Primary.Disease == "Breast Cancer")
#see what we got
summary(BCCL_Annotation)
# copy expression matrix to new dataframes
df1 <- allDepMapData$expression
df2 <- allDepMapData$copynumber
df3 <- allDepMapData$kd.ceres
df4 <- allDepMapData$kd.prob
#change column names to numbers for easier handling
names(df1) <- c(1:544)
names(df2) <- c(1:544)
names(df3) <- c(1:544)
names(df4) <- c(1:544)
#extract BCCL as dataframe
BCCL_Expression <- as.data.frame( df1[, c(BCCL_Numbers)])
BCCl_Copynumber <- as.data.frame( df2[, c(BCCL_Numbers)])
BCCL_kd.ceres <- as.data.frame( df3[, c(BCCL_Numbers)])
BCCL_kd.prob <- as.data.frame( df4[, c(BCCL_Numbers)])
dim(BCCL_Expression)
dim(BCCl_Copynumber)
dim(BCCL_kd.ceres)
dim(BCCL_kd.prob)
#reorder rows alphabetically
BCCL_Expression <- BCCL_Expression[order(rownames(BCCL_Expression)), ]
BCCl_Copynumber <- BCCl_Copynumber[order(rownames(BCCl_Copynumber)), ]
BCCL_kd.ceres <- BCCL_kd.ceres[order(rownames(BCCL_kd.ceres)), ]
BCCL_kd.prob <- BCCL_kd.prob[order(rownames(BCCL_kd.prob)), ]
#subset Mutation matrix in BCCL
BCCL_Mutation <- allDepMapData$mutation[c(BCCL_Numbers)]
```

 - include patients ID as first row in BCCL dataframes
```{r echo=FALSE}
Patients_ID <- BCCL_Annotation[1] #get patients ID from annotation matrix
tPatients_ID <- t(Patients_ID) #transpose Patients_ID because it was a column and we want to insert it as row
rownameskdc <- rownames(BCCL_kd.ceres) # save old rownames
existingDF <- BCCL_kd.ceres  # define variables for BCCL_kd.ceres
r <- 1
newrow <- tPatients_ID
#defining a function we can use more often to insert rows
insertRow <- function(existingDF, newrow, r) { 
    existingDF[seq(r+1,nrow(existingDF)+1),] <- existingDF[seq(r,nrow(existingDF)),]
    existingDF[r,] <- newrow
    existingDF
}
BCCL_kd.ceres_ID <- insertRow(existingDF, newrow, r)
rownames(BCCL_kd.ceres_ID) <- c("Deep_ID", rownameskdc) # change rownames back 
BCCL_kd.ceres_ID[1:2, ]
#do the same with other 3 dataframes
rownamesex <- rownames(BCCL_Expression) #save rownames
existingDF <- BCCL_Expression  # define variables for BCCL_Erpression
r <- 1
newrow <- tPatients_ID
BCCL_Expression_ID <- insertRow(existingDF, newrow, r)
rownames(BCCL_Expression_ID) <- c("Deep_ID", rownamesex)
rownamescn <- rownames(BCCl_Copynumber) #save rownames
existingDF <- BCCl_Copynumber  # define variables for BCCL_Copynumber
r <- 1
newrow <- tPatients_ID
BCCl_Copynumber_ID <- insertRow(existingDF, newrow, r)
rownames(BCCl_Copynumber_ID) <- c("Deep_ID", rownamescn)
rownameskdp <- rownames(BCCL_kd.prob) #save rownames
existingDF <- BCCL_kd.prob  # define variables for BCCL_kd.prob
r <- 1
newrow <- tPatients_ID
BCCL_kd.prob_ID <- insertRow(existingDF, newrow, r)
rownames(BCCL_kd.prob_ID) <- c("Deep_ID", rownameskdp)
```


PCA but not yet complete!
``` {r PCA}
install.packages("ggfortify")
library(ggfortify)

tBCCL_Expression = t(BCCL_Expression)
Expression.pca = prcomp(tBCCL_Expression, center = T, scale = F, rank = 2)  # scaling not possible due to zeros # rank = 2 since only interested in first 2 PCA components 
summary(Expression.pca)

plot(Expression.pca, type = "l")  # First 2 components = most of data  variability using Elbow-method 
autoplot(Expression.pca, label = TRUE, label.size = 3) # need to add colour based on whether driver mutation or not

Expression.pca.loadings = Expression.pca$rotation # laodings PC1 and PC2
Sor.Expr.pca.load= -Expression.pca.loadings [order(Expression.pca.loadings[,1], Expression.pca.loadings[,2]),] #managed to order PC1 column according to size (PC2 not yet)
Sor.Expr.pca.load[1:200,] # first 200 genes do not contain driver mutations - perhaps assign rank

```


Things to do: 
* manually scale Expression values 
* add column to tBCCL_expression which shows mutation or not from  MutImpact Matrix
* order matrix also according to PC2
* assign rank to genes according to PC1 und PC2 (insert 2 columns with appropriate rank- can identify driver mutations)
*k means clustering 

Same process for CERES scores (needs to be adapted if any changes made above)

```{r PCA Ceres}
tBCCL_kd.ceres = t(BCCL_kd.ceres)
Ceres.pca = prcomp(tBCCL_kd.ceres, center = T, scale = T, rank = 2)  
summary(Ceres.pca)

plot(Ceres.pca, type = "l")  # First 2 components = most of data  variability using Elbow-method 
autoplot(Ceres.pca, label = TRUE, label.size = 3) # need to add colour based on whether driver mutation or not

Ceres.pca.loadings = Ceres.pca$rotation # loadings PC1 and PC2
Sor.Ceres.pca.load= -Ceres.pca.loadings [order(Ceres.pca.loadings[,1]),] #managed to order PC1 column according to size (PC2 not yet)
Sor.Ceres.pca.load[1:200,] # first 200 genes do not contain driver mutations - perhaps assign rank


