---
title: "SST Characterisation Report"
author: "Salome"
date: "8 6 2019"
output:
  pdf_document: default
  html_document: default
---

## Verification of SST candidates by assigned k means cluster number 

After selection of the SST candidates stored in `SST_cand_x`,loops were installed to verify the coexistance of SST candidates and driver mutations in the same k means cluster. An assignment to the same cluster would confirm a possible cooperation of gene pairs due to similar CERES scores. For this, new dataframes were created  before a separate loop was installed for each driver mutation. By comparing each rowname of the `SST_cand_X` dataframes to all possible  `data_clus_3e` rownames, a coexistance of genes in the 3rd k means cluster was identified, leading to the labelling of `SST X`  in a new separate column. Otherwise no match lead to the `No SST`entry. In order to combine all four dataframes with the `rbind`command the first columns were all renamed to "p_values". FInally all confirmed SSTs were ordered alphabetically, with no gene identified twice for a different driver mutation. 

```{r}
SST_ERBB2 <- SST_cand_ERBB2 #creating new dataframes
SST_MYCBP <- SST_cand_MYCBP
SST_PARP10 <- SST_cand_PARP10
SST_PIK3CA <- SST_cand_PIK3CA

#labelling of SST candidates with "SST" or "No SST" depending on if in k means cluster 3 or not

#PARP10 candidates
SST_PARP10["SST Cluster Check"] <- sapply(1:nrow(SST_cand_PARP10), function(a){
  if (rownames(SST_cand_PARP10) %in% rownames(data_clus_3e)) {
      SST_PARP10[a,1] ="SST PARP10"
  }else{
     SST_PARP10[a,1] = "No SST"
  }
})
SST_PARP10 <- SST_PARP10 %>% rename(p_values = PARP10)

#ERBB2 candidates
SST_ERBB2["SST Cluster Check"] <- sapply(1:nrow(SST_cand_ERBB2), function(a){
  if (rownames(SST_cand_ERBB2) %in% rownames(data_clus_3e)) {
      SST_ERBB2[a,1] ="SST ERBB2"
  }else{
     SST_PARP10[a,1] = "No SST"
  }
})
SST_ERBB2 <- SST_ERBB2 %>% rename(p_values= ERBB2)

#MYCBP candidates
SST_MYCBP["SST Cluster Check"] <- sapply(1:nrow(SST_cand_MYCBP), function(a){
  if (rownames(SST_cand_MYCBP) %in% rownames(data_clus_3e)) {
      SST_MYCBP[a,1] ="SST MYCBP"
  }else{
     SST_MYCBP[a,1] = "No SST"
  }
})
SST_MYCBP <- SST_MYCBP %>% rename(p_values= MYCBP)

#PIK3CA candidates
SST_PIK3CA["SST Cluster Check"] <- sapply(1:nrow(SST_cand_PIK3CA), function(a){
  if (rownames(SST_cand_PIK3CA) %in% rownames(data_clus_3e)) {
      SST_PIK3CA[a,1] ="SST PIK3CA"
  }else{
     SST_PIK3CA[a,1] = "No SST"
  }
})
SST_PIK3CA <- SST_PIK3CA %>% rename(p_values= PIK3CA)

#All approved SSTs in one dataframe with driver mutation label and p-value
SSTs <-  do.call("rbind", list(SST_ERBB2, SST_MYCBP, SST_PARP10, SST_PIK3CA)) 
SSTs <- SSTs[order(rownames(SSTs)),] # reorder in alphabetical order 

#no SSTs the same for different driver mutations 
```

```{r SST Cluster Check}
processingfunction <- function(inputData, dtPicker, mutPicker) {
  inputData["SST Cluster Check"] <- sapply(1:nrow(dtPicker), function(a) {
    if (rownames(dtPicker) %in% rownames(data_clus_3e)) {
      inputData[a,1] <- paste0("SST", " ", mutPicker)
    } else {
      inputData[a,1] <- "No SST"  
    }
  })
  return(inputData)
}
SST_PARP10 <- processingfunction(SST_PARP10, SST_cand_PARP10, "PARP10")

```

```{r}
SST_ERBB2 <- SST_cand_ERBB2 #creating new dataframes
SST_MYCBP <- SST_cand_MYCBP
SST_PARP10 <- SST_cand_PARP10
SST_PIK3CA <- SST_cand_PIK3CA

SST_all_can <- list(SST_cand_ERBB2, SST_cand_MYCBP, SST_cand_PARP10, SST_cand_PIK3CA)# combining all dataframes into one list

processingfunction <- lapply(seq_along(SST_all_can), function(a) { # pick one dataframe fro list
  mutpicker <- SST_all_can[[a]] # define dataframe as mutpicker
  mutpicker$classifier <- NA
  evaluate <- sapply(1:nrow(mutpicker), function(b) {
    if (rownames(mutpicker)[b] %in% rownames(data_clus_3e)) {
      evaluate$paste0("SST", rownames(mutpicker)[b])
    } else {
      evaluate$paste0("No SST", rownames(mutpicker)[b]) 
    }
  evaluate <- as.vecto(evaluate) 
  mutpicker$classifier <- evaluate
  })
  return(mutpicker)
})

SST_PARP10 <- processingfunction(SST_PARP10, SST_cand_PARP10, "PARP10")
SST_PARP10 <- SST_PARP10 %>% rename(p_values=PARP10)

SST_ERBB2 <- processingfunction(SST_ERBB2, SST_cand_ERBB2, "ERBB2")
SST_ERBB2 <- SST_ERBB2 %>% rename(p_values=ERBB2)

SST_MYCBP <- processingfunction(SST_MYCBP, SST_cand_MYCBP, "MYCBP")
SST_MYCBP <- SST_MYCBP %>% rename(p_values=MYCBP)

SST_PIK3CA <- processingfunction(SST_PIK3CA, SST_cand_PIK3CA, "PIK3CA")
SST_PIK3CA <- SST_PIK3CA %>% rename(p_values=PIK3CA)

```


```{r SST Cluster Check}
SST_Data <- lapply(seq_along(SST_Data), function(a) { # pick one dataframe from list
  DvPicker <- SST_all_can[[a]] # choose DV 
  
  DvPicker$classifier <- sapply(1:nrow(DvPicker), function(b) {
    if (rownames(mutpicker)[b] %in% rownames(data_clus_3e)) {
      DvPicker[b, 5] <- "verified"
    } else {
      DvPicker[b, 5] <- "not verified" 
    }
  })
  return(mutpicker)
})
names(SST_Data) <- driver_mut
```